% !TEX root = main.tex
\chapter{LABELLED STRUCTURES AND EXPONENTIAL GENERATING FUNCTIONS}
\label{chap:labelled structures and exponential generating functions}



\section{Labelled classes}



\section{Admissible labelled constructions}



\section{Surjections, set partitions, and words}



\section{Alignments, permutations, and related structures}



\section{Labelled trees, mappings, and graphs}
% Note II.19.
\vspace{\baselineskip}
\noindent
$\rhd$ \textbf{II.19.} \textit{Labelled hierarchies.}

\vspace{\baselineskip}
\noindent
We want to derive
\begin{equation*}
    L(z) = T\left(\frac{1}{2}e^{z/2 - 1/2}\right) + \frac{z}{2} - \frac{1}{2}
\end{equation*}
from
\begin{equation*}
    L = z + e^{L} - 1 - L
\end{equation*}

\vspace{\baselineskip}
\noindent
\textbf{Step 1. Re-arranging the target equation.}
We start with the given equation, moving all terms involving $L$ to the left side and constant/independent terms to the right:
\begin{equation*}
    2L + 1 - e^{L} = z
\end{equation*}
This form does not immediately match $T(z) = ze^{T(z)}$. To make them compatible, we should attempt to isolate an exponential term
that mirrors the functional form of the tree function.

\vspace{\baselineskip}
\noindent
\textbf{Step 2. Utilizing the Lambert W identity.}
The Cayley tree function $T(z)$ is related to the \textbf{Lambert W function} (specifically $W(-z)$). The identity $T(z) = ze^{T(z)}$ can be rewritten as
\begin{equation*}
    -T(z)e^{-T(z)} = -z \leftrightarrow z = -T(-z)e^{-T(-z)}
\end{equation*}
Which implies $W(z) = -T(-z) \leftrightarrow T(z) = -W(-z)$.

\vspace{\baselineskip}
\noindent
\textbf{Step 3. Substitution and Solving.}
Let's rearrange the original equation to isolate the exponential:
\begin{equation*}
    e^{L} = 2L - z + 1
\end{equation*}
For equations where a variable appears both linearly and in the exponent, the solution is typically logarithmic or involves the $W$ function. 
Let's rewrite the equation for $L$ in the form $f(L)e^{f(L)} = \text{something}$:
\begin{equation*}
    1 = (2L - z + 1)e^{-L}
\end{equation*}
Now, we want the term inside the parentheses to match the exponent. Let $u$ be the linear part related to $L$.
If we aim for $u = \frac{z - 1}{2} - L$, then
\begin{equation*}
    2u = z - 1 - 2L = -(2L - z + 1)
\end{equation*}
Substituting this back into our equation:
\begin{equation*}
    1 = (-2u)e^{-L}
\end{equation*}
Since $u = \frac{z - 1}{2} - L$, we have $-L = u - \frac{z - 1}{2}$. Substituting this into the exponent:
\begin{equation*}
    1 = -2u \cdot e^{u - \frac{z - 1}{2}} = -2u \cdot e^{u} \cdot e^{-\frac{z - 1}{2}}
\end{equation*} 
Now, isolate $ue^{u}$:
\begin{equation*}
    ue^{u} = -\frac{1}{2}e^{\frac{z - 1}{2}}
\end{equation*}
Since $ue^{u} = v$ implies $u = W(v)$, and we know $W(v) = -T(-v)$:
\begin{equation*}
    u = W\left(-\frac{1}{2}e^{\frac{z - 1}{2}}\right) = -T\left(\frac{1}{2} e^{\frac{z - 1}{2}}\right)
\end{equation*}
Substituting $u$ back in:
\begin{equation*}
    \frac{z - 1}{2} - L = -T\left(\frac{1}{2} e^{\frac{z - 1}{2}}\right)
\end{equation*}
Solving for $L$:
\begin{equation*}
    L = T\left(\frac{1}{2} e^{\frac{z - 1}{2}}\right) + \frac{z - 1}{2}
\end{equation*}


% II.5.2.
\vspace{\baselineskip}
\noindent
\textbf{II.5.2.} \textit{\textbf{Mappings and associated graphs.}}

\vspace{\baselineskip}
\noindent
Let $\mathcal{T}$ be the class of all Cayley trees. Then
\begin{equation*}
    F = (1 - T)^{-1} \implies F_{n} = n^{n}
\end{equation*}
This result follows from Lagrange inversion. The standard formulation states that
\begin{equation*}
    z = T / \phi(T) \implies [z^{n}]T(z) = \frac{1}{n}[w^{n - 1}]\phi(w)^{n}
\end{equation*}
More generally
\begin{equation*}
    [z^{n}]H(T(z)) = \frac{1}{n}[w^{n - 1}]H'(w)\phi(w)^{n}
\end{equation*}
Substituting $H(w) = (1 - w)^{-1}$ and $\phi(w) = e^{w}$ yields
\begin{equation*}
    \begin{aligned}
        F_{n}
        &= n![z^{n}]F(z) \\
        &= n!\frac{1}{n}[w^{n - 1}]\left(\frac{1}{1 - w}\right)'(e^{w})^{n} \\
        &= (n - 1)![w^{n - 1}]\frac{e^{nw}}{(1- w)^{2}} \\
        &= (n - 1)![w^{n - 1}]\sum_{k = 0}^{\infty}(k + 1)w^{k} \cdot \sum_{j = 0}^{\infty}\frac{n^{j}}{j!}w^{j} \\
        &= (n - 1)!\sum_{j = 0}^{n - 1}((n - 1 - j) + 1)\frac{n^{j}}{j!} \\
        &= (n - 1)!\sum_{j = 0}^{n - 1}(n - j)\frac{n^{j}}{j!} \\
        &= (n - 1)!\left(\sum_{j = 0}^{n - 1}n\frac{n^{j}}{j!} - \sum_{j = 0}^{n - 1}j\frac{n^{j}}{j!}\right) \\
        &= (n - 1)!\left(\sum_{j = 0}^{n - 1}\frac{n^{j + 1}}{j!} - \sum_{j = 1}^{n - 1}\frac{n^{j}}{(j - 1)!}\right) \\
        &= (n - 1)!\left(\sum_{j = 0}^{n - 1}\frac{n^{j + 1}}{j!} - \sum_{j = 0}^{n - 2}\frac{n^{j + 1}}{j!}\right) \\
        &= (n - 1)!\frac{n^{n}}{(n - 1)!} \\
        &= n^{n}
    \end{aligned}
\end{equation*}

\vspace{\baselineskip}
\noindent
Now consider $K = \log{(1 - T)^{-1}}$. Then
\begin{equation*}
     K_{n} = n^{n - 1}Q(n)
\end{equation*}
where
\begin{equation*}
    Q_{n} = 1 + \frac{n - 1}{n} + \frac{(n - 1)(n - 2)}{n^{2}} + \cdots
\end{equation*}
From Note II.18 we have
\begin{equation*}
    n![z^{n}]\frac{T(z)^{k}}{k!} = \binom{n - 1}{k - 1}n^{n - k} \implies
        n![z^{n}]\frac{T(z)^{k}}{k} = (k - 1)!\binom{n - 1}{k - 1}n^{n - k} = \frac{(n - 1)!}{(n - k)!}n^{n - k}
\end{equation*}
Consequently
\begin{equation*}
    \begin{aligned}
        K_{n}
        &= n![z^{n}]K(z) \\
        &= n![z^{n}]\sum_{j = 1}^{\infty}\frac{T(z)^{j}}{j} \\
        &= \sum_{j = 1}^{\infty}n![z^{n}]\frac{T(z)^{j}}{j} \\
        &= \sum_{j = 1}^{\infty}\frac{(n - 1)!}{(n - j)!}n^{n - j} \\
        &= n^{n - 1}\sum_{j = 1}^{\infty}\frac{(n - 1)!}{(n - j)!n^{j - 1}} \\
        &= n^{n - 1}\left(1 + \frac{n - 1}{n} + \frac{(n - 1)(n - 2)}{n^{2}} + \cdots\right)
    \end{aligned}
\end{equation*}

\vspace{\baselineskip}
\noindent
\textbf{II.5.2.} \textit{\textbf{Constrained mappings.}}

\vspace{\baselineskip}
\noindent
The class of all maps without fixed points has EGF:
\begin{equation*}
    \frac{e^{-T(z)}}{1 - T(z)}
\end{equation*}
We aim to show that
\begin{equation*}
    n![z^{n}]\frac{e^{-T(z)}}{1 - T(z)} = (n - 1)^{n}
\end{equation*}
Substituting $H(w) = \frac{e^{-w}}{1 - w}$ and $\phi(w) = e^{w}$ into Lagrange inversion yields
\begin{equation*}
    \begin{aligned}
        n![z^{n}]\frac{e^{-T(z)}}{1 - T(z)}
        &= n!\frac{1}{n}[w^{n - 1}]\left(\frac{e^{-w}}{1 - w}\right)'(e^{w})^{n} \\
        &= (n - 1)![w^{n - 1}]\frac{we^{(n - 1)w}}{(1 - w)^{2}} \\
        &= (n - 1)![w^{n - 2}]\frac{e^{(n - 1)w}}{(1 - w)^{2}} \\
        &= (n - 1)![w^{n - 2}]\sum_{k = 0}^{\infty}(k + 1)w^{k} \cdot \sum_{j = 0}^{\infty}\frac{(n - 1)^{j}}{j!}w^{j} \\
        &= (n - 1)!\sum_{j = 0}^{n - 2}((n - 2 - j) + 1)\frac{(n - 1)^{j}}{j!} \\
        &= (n - 1)!\sum_{j = 0}^{m - 1}(m - j)\frac{m^{j}}{j!} \\
        &= (n - 1)!\frac{m^{m}}{(m - 1)!} \\
        &= (n - 1)!\frac{(n - 1)^{n - 1}}{(n - 2)!} \\
        &= (n - 1)^{n}
    \end{aligned}
\end{equation*}


% II.5.3.
\vspace{\baselineskip}
\noindent
\textbf{II.5.3.} \textit{\textbf{Unrooted trees and acyclic graphs.}}

\vspace{\baselineskip}
\noindent
To understand why $U(z) = \int_{0}^{z}T(w)\frac{dw}{w}$, we begin with the exponential generating functions:
\begin{equation*}
    T(z) = \sum_{n \geq 0}T_{n}\frac{z^{n}}{n!}, \quad
    U(z) = \sum_{n \geq 0}U_{n}\frac{z^{n}}{n!}
\end{equation*}
Differentiate $U(z)$ and then multiply by $z$ (noting that $T_{0} = 0$) gives
\begin{equation*}
    z\frac{d}{dz}U(z) = z\frac{d}{dz}\sum_{n \geq 0}U_{n}\frac{z^{n}}{n!} = z\sum_{n \geq 1}nU_{n}\frac{z^{n - 1}}{n!}
        = \sum_{n \geq 1}T_{n}\frac{z^{n}}{n!} = T(z)
\end{equation*}
Rearranging yields $\frac{d}{dz}U(z) = \frac{T(z)}{z}$. Integrating both sides with respect to $z$ and using $w$ as a
dummy variable for the integration from $0$ to $z$, we obtain
\begin{equation*}
    U(z) - U(0) = \int_{0}^{z}\frac{T(w)}{w}dw
\end{equation*}
The EGF for rooted trees, $T(z)$, satisfies the functional equation:
\begin{equation*}
    T(z) = ze^{T(z)}
\end{equation*}
which can be rearranged to $z = Te^{-T}$. Differentiating with respect to $T$ gives
\begin{equation*}
    dz = \left(e^{-T} - Te^{-T}\right)dT = e^{-T}(1 - T)dT
\end{equation*}
Substituting $z = Te^{-T}$ and $dz$ into the integral $U(z) = \int\frac{T(z)}{z}dz$ yields
\begin{equation*}
    U(z) = \int \frac{T}{Te^{-T}} \cdot e^{-T}(1 - T)dT = \int (1 - T)dT = T - \frac{T^{2}}{2} + C
\end{equation*}
Since $U(0) = 0$ and $T(0) = 0$, the constant $C = 0$.


% Note II.22.
\vspace{\baselineskip}
\noindent
$\rhd$ \textbf{II.22.} \textit{2–Regular graphs.}

\vspace{\baselineskip}
\noindent
We establish a bijection between clouds and 2-regular graphs on $n$ labeled vertices. Number the lines $1, 2, ..., n$.
\begin{itemize}
    \item Line $i$ in the cloud $\leftrightarrow$ vertex $i$ in the graph.
    \item Intersection point $\{i,\; j\}$ in the cloud $\leftrightarrow$ edge $i - j$ in the graph.
\end{itemize}
Let $n_{i}$ be the number of chosen intersection points on line $i$.
Since each point lies on two lines, $\sum_{i = 1}^{n}n_{i} = 2n$.
\textit{If no three points are collinear}, then $n_{i} \leq 2$ for all $i$, so $\sum_{i = 1}^{n}n_{i} \leq 2n$.
Equality $\sum_{i = 1}^{n}n_{i} = 2n$ forces $n_{i} = 2$ for every $i$, i.e. every vertex has degree 2 in the graph.

\vspace{\baselineskip}
\noindent
Conversely, if every vertex has degree 2, then $n_{i} = 2$ for all $i$, so each line contains exactly two intersection points — hence no three collinear.

\vspace{\baselineskip}
\noindent
Thus the cloud condition and the 2-regular graph condition are equivalent. 


% Note II.23.
\vspace{\baselineskip}
\noindent
$\rhd$ \textbf{II.23.} \textit{Wright’s surgery.}

\vspace{\baselineskip}
\noindent
\textbf{The Surgery in Forward Direction: Deconstruction}

\noindent
We start with a connected graph $G$ of excess $k+1$. Pick an edge of $G$ and an orientation (a direction) for it.
This gives us a starting vertex $A$ and an ending vertex $B$. Remove this directed edge from the graph.
Depending on the role that edge played in the connectivity of $G$, one of two mutually exclusive cases occurs
\begin{itemize}
    \item \textbf{Case 1 (Connected)}: The removed edge was part of a cycle. The graph remains connected. Because we lost an edge but kept all vertices, 
        the excess drops to $k$. The graph now possesses two designated, ordered, pointed vertices: a ``source'' point at $A$ and a ``target'' point at $B$.
    \item \textbf{Case 2 (Disconnected)}: The removed edge was a bridge. The graph splits into two separate connected components, $G_{1}$ and $G_{2}$.
        If $G_{1}$ takes on an excess of $h$, then $G_{2}$ must take on the remaining excess of $k - h$. Furthermore, $G_{1}$ contains exactly one
        pointed vertex (where $A$ used to connect), and $G_{2}$ contains exactly one pointed vertex (where $B$ used to connect).
\end{itemize}

\vspace{\baselineskip}
\noindent
\textbf{The Surgery in Backward Direction: Reconstruction}

\noindent
We want to reverse this process to build all possible directed-edge-pointed graphs of excess $k+1$. We do this by reversing both cases
\begin{itemize}
    \item \textbf{Reversing Case 1}: Take a single connected graph of excess $k$. Pick any two distinct vertices in an ordered fashion, call them $A$ and $B$,
        and draw a directed edge from $A$ to $B$. If $A$ and $B$ already had an edge between them, adding a new one creates an illegal multi-edge.
        We must forbid picking pairs of vertices that are already directly connected.
    \item \textbf{Reversing Case 2}: Take two completely independent connected components — one of excess $h$ and one of excess $k - h$.
        Pick one vertex in the first component ($A$) and one vertex in the second component ($B$). Draw a directed edge from $A$ to $B$,
        merging them into a single connected graph of excess $k + 1$.
\end{itemize}

\noindent
Now let's align these actions with the differential recurrence:
\begin{equation*}
    2\partial_{y}w_{k + 1} = \left(z^{2}\partial_{z}^{2}w_{k} - 2y\partial_{y}w_{k}\right) +
        \sum_{h = -1}^{k + 1}\left(z\partial_{z}w_{h}\right) \cdot \left(z\partial_{z}w_{k - h}\right)
\end{equation*}

\noindent
Left-Hand Side: The Initial Pool
\begin{itemize}
    \item $w_{k + 1}$: Represents the universe of connected graphs of excess $k + 1$.
    \item $\partial_{y}$: Differentiates with respect to the edge-marker $y$. Combinatorially, $\frac{\partial}{\partial y}(y^{E}) = E \cdot y^{E - 1}$,
        which means ``choose one of the $E$ edges and remove it''.
    \item $2$: Accounts for the orientation of the removed edge. Since an unoriented edge has 2 endpoints, choosing a direction means deciding
        which endpoint is the source ($A$) and which is the target ($B$).
    \item Together ($2\partial_{y}w_{k + 1}$): Represents a graph of excess $k + 1$ with one structurally deleted, but memory-tracked, directed edge.
\end{itemize}

\noindent
Right-Hand Side, Term 1: The Connected Outcome (Case 1)
\begin{itemize}
    \item $w_{k}$: The base graph of excess $k$.
    \item $\partial_{z}^{2}$: Differentiates with respect to the vertex-marker $z$ twice. This selects an ordered pair of distinct vertices ($A$ and $B$).
    \item $z^{2}$: Multiplies $z^{2}$ back in, ensuring that the total vertex count of the graph remains unchanged after the selection.
    \item $-2y\partial_{y}w_{k}$: The exclusion rule. $y\partial_{y}w_{k}$ selects an existing edge in the excess-$k$ graph.
        That existing edge has 2 possible orientations (endpoints ordered as $(A,B)$ or $(B,A)$). We subtract this out to prevent
        the $z^{2}\partial_{z}^{2}$ operator from choosing an existing edge's endpoints and accidentally creating a double-edge.
\end{itemize}

\noindent
Right-Hand Side, Term 2: The Disconnected Outcome (Case 2)
\begin{itemize}
    \item $w_{h}$ and $w_{k - h}$: The two separate, independent components. Their generating functions are multiplied together because
        the components are combinatorially disjoint.
    \item $z\partial_{z}$: The pointing operator for vertices ($z\frac{\partial}{\partial z}(z^{V}) = V \cdot z^{V}$).
        \begin{itemize}
            \item Applied to the first component ($z\partial_{z}w_{h}$), it selects exactly one vertex ($A$) to act as the source bridgehead.
            \item Applied to the second component ($z\partial_{z}w_{k - h}$), it selects exactly one vertex ($B$) to act as the target bridgehead.
        \end{itemize}
    \item $\sum_{h = -1}^{k + 1}$: Sums over all valid ways the total remaining excess $k$ can be distributed between the two components.
        Because the simplest possible connected graph is a tree (which has an excess of $E - V = -1$), the lowest possible excess a component can hold is $-1$.
\end{itemize}



\section{Additional constructions}
% Note II.28.
\vspace{\baselineskip}
\noindent
$\rhd$ \textbf{II.28.} \textit{Do two permutations generate the symmetric group?}

\vspace{\baselineskip}
\noindent
We want to show
\begin{equation*}
    n!^{2}\pi_{n} = (n - 1)!I_{n + 1}
\end{equation*}
The EGF of pairs of permutations $\sum_{n \geq 0}n!^{2}\frac{z^{n}}{n!} = \sum_{n \geq 0}n!z^{n}$ happens to be the OGF $P(z)$ of all permutations.
By the exponential formula, the EGF $C(z)$ of connected graphs satisfies
\begin{equation*}
    C(z) = \log{P(z)} = \log{\left(\sum_{n \geq 0}n!z^{n}\right)}
\end{equation*}
Since every permutation can be uniquely decomposed into a sequence of indecomposable components, the OGF $I(z)$ of indecomposable permutations satisfies
\begin{equation*}
    I(z) = 1 - \frac{1}{P(z)}
\end{equation*}
Let's find a differential relationship. Take the derivative of $C(z)$ with respect to $z$:
\begin{equation*}
    C'(z) = \frac{P'(z)}{P(z)} = P'(z)(1 - I(z)) = P'(z) - P'(z)I(z)
\end{equation*}
Now, let's extract the coefficient of $z^{n - 1}$ from both sides of this derivative equation.

\vspace{\baselineskip}
\noindent
\textbf{Left-Hand Side}
Since $C(z) = \sum_{n \geq 1}(n!^{2}\pi_{n})\frac{z^{n}}{n!}$, its derivative is
\begin{equation*}
    C'(z) = \sum_{n \geq 1}(n!^{2}\pi_{n})\frac{z^{n - 1}}{(n - 1)!}
\end{equation*}
Thus, the coefficient of $z^{n - 1}$ is
\begin{equation*}
    [z^{n - 1}]C'(z) = \frac{n!^{2}\pi_{n}}{(n - 1)!} = n \cdot n!\pi_{n}
\end{equation*}

\vspace{\baselineskip}
\noindent
\textbf{Right-Hand Side}
Differentiate both sides of $P(z)(1 - I(z)) = 1$ with respect to $z$:
\begin{equation*}
    P'(z)(1 - I(z)) - P(z)I'(z) = 0 \implies P'(z) - P'(z)I(z) = P(z)I'(z)
\end{equation*}
Now we just need to extract the coefficient of $z^{n - 1}$ from $P(z)I'(z)$:
\begin{equation*}
    [z^{n - 1}](P'(z) - P'(z)I(z)) = [z^{n - 1}]P(z)I'(z) = \sum_{k = 1}^{n}kI_{k}(n - k)!
\end{equation*}
Let's denote this sum by $S_{n}$. We will show $S_{n} = I_{n + 1}$.
Extracting the coefficient of $z^{N}$ from $P(z) = 1 + I(z)P(z)$ for $N \geq 1$ yields the recurrence (noting that $I_{0} = 0$):
\begin{equation} \label{eq:II.28.1}
    \sum_{k = 1}^{N}I_{k}(N - k)! = N!
\end{equation}
Use \ref{eq:II.28.1} with $N = n + 1$:
\begin{equation} \label{eq:II.28.2}
    \sum_{k = 1}^{n + 1}I_{k}(n + 1 - k)! = (n + 1)! \implies \sum_{k = 1}^{n}I_{k}(n + 1 - k)! = (n + 1)! - I_{n + 1}
\end{equation}
Expand the left hand side of \ref{eq:II.28.2}:
\begin{equation*}
    \begin{aligned}
        \sum_{k = 1}^{n}I_{k}(n + 1 - k)!
        &= \sum_{k = 1}^{n}I_{k}(n + 1 - k)(n - k)! \\
        &= (n + 1)\sum_{k = 1}^{n}I_{k}(n - k)! - \sum_{k = 1}^{n}kI_{k}(n - k)! \\
    \end{aligned}
\end{equation*}
By \ref{eq:II.28.1} with $N = n$:
\begin{equation*}
    \sum_{k = 1}^{n}I_{k}(n - k)! = n!
\end{equation*}
Hence the left hand side of \ref{eq:II.28.2} equals
\begin{equation*}
    \sum_{k = 1}^{n}I_{k}(n + 1 - k)! = (n + 1)n! - S_{n} = (n + 1)! - S_{n}
\end{equation*}
Substituting into \ref{eq:II.28.2} gives $S_{n} = I_{n + 1}$.

\vspace{\baselineskip}
\noindent
Finally $n \cdot n!\pi_{n} = S_{n} \implies n!^{2}\pi_{n} = (n - 1)!I_{n + 1}$.



\section{Perspective}