% !TEX root = main.tex
\chapter{COMBINATORIAL PARAMETERS AND MULTIVARIATE GENERATING FUNCTIONS}
\label{chap:combinatorial parameters and multivariate generating functions}

\section{An introduction to bivariate generating functions (BGFs)}

\section{Bivariate generating functions and probability distributions}

\section{Inherited parameters and ordinary MGFs}
% Note III.7.
\noteblock{III.7}{Largest summands in compositions.}

Let $M$ be the largest summand in a random composition. We claim that for any $\epsilon \in (0, 1)$,
\begin{equation*}
    \mathbb{P}\left((1 - \epsilon)\log_{2}n \leq M \leq (1 + \epsilon)\log_{2}n\right) \to 1
\end{equation*}

\stephead{1}{Upper bound via the first moment method.}

Set $r_{+} = \lceil(1 + \epsilon)\log_{2}n\rceil$. If $M \geq r_{+}$, then some summand has size at least $r_{+}$.
Let $S = \sum_{k = r_{+}}^{n}X_{k}$, where $X_{k}$ counts summands of size exactly $k$.
Then $S$ counts summands of size at least $r_+$, so
\begin{equation*}
    \mathbb{P}(M \geq r_{+}) = \mathbb{P}(S \geq 1) \leq \frac{\mathbb{E}[S]}{1} = \sum_{k = r_{+}}^{n}\mathbb{E}[X_{k}]
\end{equation*}
Use the mean estimate from Proposition III.4 in its explicit form. For $1 \leq k \leq n$,
\begin{equation*}
    \mathbb{E}[X_k] = \frac{n-k+3}{2^{k+1}} \leq \frac{n+2}{2^{k+1}}.
\end{equation*}
For all large enough $n$, this is at most $\frac{n}{2^k}$, which is more than enough here.
Thus
\begin{equation*}
    \mathbb{E}[S] \leq \sum_{k = r_{+}}^{n} \frac{n}{2^{k}} < \frac{n}{2^{r_{+} - 1}}
        \leq \frac{2n}{2^{(1 + \epsilon)\log_{2}n}} = 2n^{-\epsilon} \to 0
\end{equation*}
$M > (1 + \epsilon)\log_{2}n$ implies $M \geq \lceil (1 + \epsilon)\log_{2}n \rceil = r_{+}$, so
\begin{equation*}
    \mathbb{P}(M > (1 + \epsilon)\log_{2}n) \leq \mathbb{P}(M \geq r_{+}) \to 0
\end{equation*}

\stephead{2}{Lower bound via the second moment method.}

For the lower bound, view a composition of $n$ as $n - 1$ independent gaps, each cut with probability $1/2$.
Let $r_{-} = \lceil (1 - \epsilon)\log_{2}n \rceil$. Split the $n$ ones into
$m = \left\lfloor \frac{n}{r_{-}} \right\rfloor$ disjoint blocks of length $r_{-}$, ignoring the leftovers.
Each block has $r_{-} - 1$ internal gaps, so it contains no internal cut with probability
$p = \left(\frac{1}{2} \right)^{r_{-} - 1}$.

Define the indicator random variables
\begin{equation*}
    I_{i} = \begin{cases}
        1 & \text{if block} \; $i$ \; \text{has no internal cuts} \\
        0 & \text{otherwise}
    \end{cases}
\end{equation*}
for $i = 1, \cdots, m$. These indicators depend on disjoint sets of gaps, so they are
independent Bernoulli variables with common parameter $p$. Let $Y = \sum_{i = 1}^{m} I_{i}$,
the number of blocks that survive intact.
If $Y \geq 1$, then at least one block is contained in a summand of size at least $r_{-}$;
the summand may extend beyond the block, but that only helps. Therefore
$\mathbb{P}(M \geq r_{-}) \geq \mathbb{P}(Y > 0)$.

Now the Paley–Zygmund inequality does exactly the required job:
\begin{equation*}
    \mathbb{P}(Y > 0) \geq \frac{(\mathbb{E}[Y])^{2}}{\mathbb{E}[Y^{2}]}
\end{equation*}
In our setting, $Y \sim \text{Binomial}(m, p)$.

First compute the expectation:
\begin{equation*}
    \mathbb{E}[Y] = mp = \left\lfloor \frac{n}{r_{-}} \right\rfloor \left(\frac{1}{2} \right)^{r_{-} - 1}
\end{equation*}
Let $a=(1-\epsilon)\log_2 n$. Since $r_-=\lceil a\rceil$, we have $r_- -1 < a$.
Also $r_-\leq a+1$, so for all large enough $n$, $a\geq 1$ and $n\geq 4a$, giving
\begin{equation*}
    \left\lfloor \frac{n}{r_-}\right\rfloor
        \geq \frac{n}{r_-}-1
        \geq \frac{n}{a+1}-1
        \geq \frac{n}{2a}-1
        \geq \frac{n}{4a}.
\end{equation*}
Also,
\begin{equation*}
    \left(\frac{1}{2}\right)^{r_- - 1}
        > \left(\frac{1}{2}\right)^a
        = n^{-(1-\epsilon)}.
\end{equation*}
Putting these together,
\begin{equation*}
    \mathbb{E}[Y] > \frac{n^{\epsilon}}{4(1 - \epsilon)\log_{2}n} \to \infty \quad \text{as } n \to \infty
\end{equation*}

Next control the second moment. Since $Y$ is binomial,
\begin{equation*}
    \text{Var}(Y) = mp(1 - p) \leq mp = \mathbb{E}[Y]
\end{equation*}

Substitute this into Paley–Zygmund:
\begin{equation*}
    \mathbb{P}(Y > 0) \geq \frac{(\mathbb{E}[Y])^{2}}{\mathbb{E}[Y^{2}]} = \frac{(\mathbb{E}[Y])^{2}}{\text{Var}(Y) + (\mathbb{E}[Y])^{2}}
            \geq \frac{(\mathbb{E}[Y])^{2}}{\mathbb{E}[Y] + (\mathbb{E}[Y])^{2}} = \frac{1}{1 + \frac{1}{\mathbb{E}[Y]}} \to 1
\end{equation*}
Finally, $M \geq r_{-}$ implies $M \geq (1 - \epsilon)\log_{2}n$, so
\begin{equation*}
    \mathbb{P}(M \geq (1 - \epsilon)\log_{2}n) \geq \mathbb{P}(M \geq r_{-}) \geq \mathbb{P}(Y > 0) \to 1
\end{equation*}
Together with the upper-tail estimate, this proves the claim, since
\begin{equation*}
    \mathbb{P}\left(M < (1-\epsilon)\log_2 n \text{ or } M > (1+\epsilon)\log_2 n\right) \to 0.
\end{equation*}

\stephead{3}{Direct proof of the Paley--Zygmund bound.}
\begin{equation*}
    \mathbb{P}(Y > 0) \geq \frac{(\mathbb{E}[Y])^{2}}{\mathbb{E}[Y^{2}]}
\end{equation*}
Start from $\mathbb{E}[Y]$ and split according to whether $Y > 0$:
\begin{equation*}
    \mathbb{E}[Y] = \mathbb{E}[Y \cdot \mathbf{1}_{Y > 0}] + \mathbb{E}[Y \cdot \mathbf{1}_{Y = 0}] = \mathbb{E}[Y \cdot \mathbf{1}_{Y > 0}]
\end{equation*}
Now apply Cauchy–Schwarz:
\begin{equation*}
    \mathbb{E}[Y] \leq \sqrt{\mathbb{E}[Y^{2}]} \sqrt{\mathbb{E}[\mathbf{1}_{Y > 0}]} = \sqrt{\mathbb{E}[Y^{2}]} \sqrt{\mathbb{P}(Y > 0)}
\end{equation*}
Since everything is nonnegative, we may square both sides:
\begin{equation*}
    (\mathbb{E}[Y])^{2} \leq \mathbb{E}[Y^{2}] \cdot \mathbb{P}(Y > 0) \implies \mathbb{P}(Y > 0) \geq \frac{(\mathbb{E}[Y])^{2}}{\mathbb{E}[Y^{2}]}
\end{equation*}

\section{Inherited parameters and exponential MGFs}
% Note III.10.
\noteblock{III.10}{A hundred prisoners II.}

The randomisation trick is easiest to see in a toy version with $10$ prisoners, where each
prisoner may open $5$ drawers. Suppose a malicious director puts the cards in one long cycle:
physical drawer $i$ contains card $i+1$, with drawer $10$ containing card $1$. If the prisoners
use the usual cycle-following strategy directly, they all follow this $10$-cycle, so the strategy
fails.

Before the game, however, the prisoners may secretly randomise the drawer names. For example,
let virtual drawer labels correspond to physical drawers as follows:
\begin{equation*}
    \rho =
    \begin{pmatrix}
        1&2&3&4&5&6&7&8&9&10\\
        7&3&9&1&8&2&10&4&6&5
    \end{pmatrix}.
\end{equation*}
When a prisoner says ``open drawer $j$'', he actually opens physical drawer $\rho(j)$.
The card found there then gives the next virtual drawer label to try.

In this example the effective permutation followed by the prisoners is
\begin{equation*}
    j \longmapsto \sigma(\rho(j)),
\end{equation*}
where $\sigma$ is the director's physical arrangement. Its cycle decomposition is
\begin{equation*}
    (1\,8\,5\,9\,7)(2\,4)(3\,10\,6).
\end{equation*}
All cycles now have length at most $5$, so the cycle-following strategy succeeds, even though
the director originally chose the worst possible $10$-cycle.

The point is not this particular lucky relabelling. If $\rho$ is chosen uniformly at random and
kept secret, then for any fixed director's arrangement $\sigma$, the product $\sigma\circ\rho$
is uniformly distributed over all permutations. Thus randomising the drawer names reduces the
malicious case back to the ordinary random-permutation calculation.

% Note III.15.
\noteblock{III.15}{Distinct component sizes in set partitions.}

The parameter is not the number of components, but the number of \textit{sizes} that occur.
So we group components by their size. For each fixed $n$, there are only two possibilities:
use no component of size $n$, or use a non-empty set of such components and pay one marker
$u$ for the fact that size $n$ appears.

For set partitions, components of size $n$ are blocks $\mathrm{S}\textsc{et}_{n}(\mathcal{Z})$,
whose EGF is $z^n/n!$. Thus the contribution of one fixed size $n$ is
\begin{equation*}
    1 + u\left(\exp\left(\frac{z^n}{n!}\right) - 1\right)
        = 1 - u + u\exp\left(\frac{z^n}{n!}\right).
\end{equation*}
Multiplying over all possible block sizes gives the BGF for the number of distinct block sizes:
\begin{equation*}
    \prod_{n \geq 1}\left(1 - u + u\exp\left(\frac{z^n}{n!}\right)\right).
\end{equation*}

For permutations the same argument applies, with cycles replacing blocks. A cycle of length
$n$ has EGF $z^n/n$, so the fixed-size contribution is
\begin{equation*}
    1 + u\left(\exp\left(\frac{z^n}{n}\right) - 1\right)
        = 1 - u + u\exp\left(\frac{z^n}{n}\right).
\end{equation*}
Hence the BGF for the number of distinct cycle sizes is
\begin{equation*}
    \prod_{n \geq 1}\left(1 - u + u\exp\left(\frac{z^n}{n}\right)\right).
\end{equation*}

\section{Recursive parameters}
% Example III.14.
\exampleblock{III.14}{Leaves and node types in binary trees.}

This is the pruned-binary-tree class from Example I.13: start with an ordinary binary tree and
remove the external nodes. The book writes $B$ here, but to avoid confusing it with ordinary
binary trees, we write it as $\bar{B}$. Thus $\bar{B}$ is the class of non-empty plane trees
where each node has $0$, $1$, or $2$ ordered children. Its OGF is
\begin{equation*}
    \bar{B}(z) = C(z) - 1 = \frac{1 - \sqrt{1 - 4z}}{2z} - 1 = \frac{1 - 2z - \sqrt{1 - 4z}}{2z}
\end{equation*}
where $C(z)$ is the usual Catalan generating function, satisfying $C(z) = 1 + zC(z)^{2}$.
Equivalently,
\begin{equation*}
    \bar{B} = \mathcal{Z} + 2\mathcal{Z}\bar{B} + \mathcal{Z}\bar{B}^{2},
    \qquad
    \bar{B}(z) = z + 2z\bar{B}(z) + z\bar{B}(z)^2.
\end{equation*}
The factor $2$ records whether the unique child of a $1$-node is on the left or on the right.

The book marks the three node types at once:
\begin{equation*}
    \bar{B}(z,u_0,u_1,u_2)
        = zu_0 + 2zu_1\bar{B}(z,u_0,u_1,u_2)
            + zu_2\bar{B}(z,u_0,u_1,u_2)^2.
\end{equation*}
Lagrange inversion then gives
\begin{equation*}
    [z^n u_0^{k_0}u_1^{k_1}u_2^{k_2}]\bar{B}(z,u_0,u_1,u_2)
    =
    \frac{1}{n}\binom{n}{k_0,k_1,k_2}2^{k_1},
\end{equation*}
under the natural constraints $k_0+k_1+k_2=n$ and $k_0=k_2+1$. Below we rederive the
three first moments directly, one marker at a time.

\subhead{Leaves}
\begin{equation*}
    \bar{B}(z, u) = zu + 2z\bar{B}(z, u) + z\bar{B}(z, u)^{2} \implies
        \frac{\partial \bar{B}}{\partial u} = z + 2z \frac{\partial \bar{B}}{\partial u} + 2z\bar{B} \frac{\partial \bar{B}}{\partial u} \implies
\end{equation*}
\begin{equation*}
    \frac{\partial \bar{B}}{\partial u} = \frac{z}{1 - 2z - 2z\bar{B}} \implies
        \left.\frac{\partial \bar{B}}{\partial u}\right|_{u = 1} = \frac{z}{1 - 2z - 2z\bar{B}(z)} = \frac{z}{\sqrt{1 - 4z}} \implies
\end{equation*}
\begin{equation*}
    [z^{n}] \left.\frac{\partial \bar{B}}{\partial u}\right|_{u = 1} = [z^{n - 1}] (1 - 4z)^{-1/2} = \binom{2n - 2}{n - 1} = \frac{(2n - 2)!}{(n - 1)!^{2}}
\end{equation*}

\subhead{1-nodes}
\begin{equation*}
    \bar{B}(z, u) = z + 2zu\bar{B}(z, u) + z\bar{B}(z, u)^{2} \implies
        \frac{\partial \bar{B}}{\partial u} = 2z\bar{B} + 2zu \frac{\partial \bar{B}}{\partial u} + 2z\bar{B} \frac{\partial \bar{B}}{\partial u} \implies
\end{equation*}
\begin{equation*}
    \frac{\partial \bar{B}}{\partial u} = \frac{2z\bar{B}}{1 - 2zu - 2z\bar{B}} \implies
        \left.\frac{\partial \bar{B}}{\partial u}\right|_{u = 1} = \frac{2z\bar{B}(z)}{1 - 2z - 2z\bar{B}(z)} = \frac{1 - 2z}{\sqrt{1 - 4z}} - 1 \implies
\end{equation*}
\begin{equation*}
    [z^{n}] \left.\frac{\partial \bar{B}}{\partial u}\right|_{u = 1} = [z^{n}] \left((1 - 4z)^{-1/2} - 2 \frac{z}{\sqrt{1 - 4z}}\right)
        = \binom{2n}{n} - 2 \binom{2n - 2}{n - 1}
\end{equation*}

\subhead{2-nodes}
\begin{equation*}
    \bar{B}(z, u) = z + 2z\bar{B}(z, u) + zu\bar{B}(z, u)^{2} \implies
        \frac{\partial \bar{B}}{\partial u} = 2z \frac{\partial \bar{B}}{\partial u} + z\bar{B}^{2} + 2zu\bar{B} \frac{\partial \bar{B}}{\partial u} \implies
\end{equation*}
\begin{equation*}
    \frac{\partial \bar{B}}{\partial u} = \frac{z\bar{B}^{2}}{1 - 2z - 2zu\bar{B}} \implies
        \left.\frac{\partial \bar{B}}{\partial u}\right|_{u = 1} = \frac{z\bar{B}(z)^{2}}{1 - 2z - 2z\bar{B}(z)}
        = \frac{z\bar{B}(z)^{2}}{\sqrt{1 - 4z}} = \frac{z^{3}C(z)^{4}}{\sqrt{1 - 4z}}
\end{equation*}
At this point a standard Catalan identity saves a fair amount of algebra. Since
$C'(z)$ satisfies $C'(z) = C(z)^{2} + 2zC(z)C'(z)$, we have
\begin{equation*}
    C'(z) = \frac{C(z)^{2}}{1 - 2zC(z)} = \frac{C(z)^{2}}{\sqrt{1 - 4z}}
\end{equation*}
Plugging that back in gives:
\begin{equation*}
    \left.\frac{\partial \bar{B}}{\partial u}\right|_{u = 1} = z^{3}C(z)^{2} \cdot \frac{C(z)^{2}}{\sqrt{1 - 4z}}
        = z^{3}C(z)^{2}C'(z) = \frac{z^{3}}{3} \frac{\dd}{\dd z} C(z)^{3} = z^{3} \frac{\dd}{\dd z} F(z)
\end{equation*}
where $F(z) = \frac{1}{3} C(z)^{3}$.
Now:
\begin{equation*}
    \begin{aligned}
        [z^{n}] \left.\frac{\partial \bar{B}}{\partial u}\right|_{u = 1}
        &= [z^{n}] z^{3} \frac{\dd}{\dd z} F(z) \\
        &= [z^{n - 3}] \frac{\dd}{\dd z} F(z) \\
        &= (n - 2)F_{n - 2} \\
        &= \frac{n - 2}{3} [z^{n - 2}] C(z)^{3} \\
        &= \frac{n - 2}{3} \cdot \frac{3}{n + 1} \binom{2n - 2}{n - 2} \\
        &= \binom{2n - 2}{n - 3}
    \end{aligned}
\end{equation*}

After collecting the three contributions, the total number of leaves, $1$-nodes, and $2$-nodes
over all trees of size $n$ is
\begin{equation*}
    \binom{2n - 2}{n - 1} + \binom{2n}{n} - 2 \binom{2n - 2}{n - 1} + \binom{2n - 2}{n - 3} = \frac{n}{n + 1} \binom{2n}{n}
\end{equation*}
which is exactly $n\bar{B}_n$, as it should be.

Dividing by $\bar{B}_n = \frac{1}{n+1}\binom{2n}{n}$ gives the exact means:
\begin{equation*}
    \mathbb{E}(\text{leaves})
        = \frac{n(n+1)}{2(2n-1)} \sim \frac{n}{4},
\end{equation*}
\begin{equation*}
    \mathbb{E}(1\text{-nodes})
        = \frac{(n+1)(n-1)}{2n-1} \sim \frac{n}{2},
\end{equation*}
\begin{equation*}
    \mathbb{E}(2\text{-nodes})
        = \frac{(n-1)(n-2)}{2(2n-1)} \sim \frac{n}{4}.
\end{equation*}
Thus leaves, left-branching nodes, right-branching nodes, and $2$-nodes all have asymptotic
proportion $1/4$, since the $1$-nodes split equally between left and right.

% Note III.18.
\noteblock{III.18}{Node-degree profile in simple varieties of trees.}
For a simple variety of trees,
\begin{equation*}
    T(z) = z\phi(T(z)), \qquad \phi(w)=\sum_{j\geq 0}\phi_j w^j.
\end{equation*}
To mark nodes with exactly $k$ children, replace the single term $\phi_kT^k$ by
$u\phi_kT^k$. Thus
\begin{equation*}
    T(z, u) = z\left(\phi(T(z, u)) + \phi_{k}(u - 1)T(z, u)^{k}\right)
\end{equation*}
Let
\begin{equation*}
    \Omega(z) = \left.\frac{\partial}{\partial u}T(z,u)\right|_{u=1},
\end{equation*}
the cumulative GF for the number of degree-$k$ nodes. Differentiating with respect to $u$
and then setting $u=1$ gives:
\begin{equation*}
    \frac{\partial T}{\partial u} =
        z\left(\phi'(T)\frac{\partial T}{\partial u} + \phi_{k}T^{k} + k\phi_{k}(u - 1)T^{k - 1} \frac{\partial T}{\partial u} \right) \implies
\end{equation*}
\begin{equation*}
    \Omega(z) = z\left(\phi'(T(z))\Omega(z) + \phi_{k}T(z)^{k}\right) \implies
        \Omega(z) = z \frac{\phi_{k}T(z)^{k}}{1 - z\phi'(T(z))}
\end{equation*}
Since $T(z) = z\phi(T(z))$, differentiating with respect to $z$ yields:
\begin{equation*}
    T'(z) = \phi(T(z)) + z\phi'(T(z))T'(z) \implies 1 = \frac{\phi(T(z))}{T'(z)} + z\phi'(T(z))
\end{equation*}
Hence,
\begin{equation*}
    \Omega(z) = \phi_{k}zT(z)^{k} \cdot \frac{T'(z)}{\phi(T(z))} = \phi_{k}zT(z)^{k} \cdot \frac{T'(z)}{T(z)/z} = \phi_{k}z^{2}T(z)^{k - 1}T'(z)
\end{equation*}
For Cayley trees, $\phi(w)=e^w$, so $\phi_k=1/k!$ and $T(z)=ze^{T(z)}$. Then
\begin{equation*}
    \begin{aligned}
        n![z^{n}] \Omega(z)
        &= n![z^{n}] \frac{1}{k!} z^{2} \frac{\dd}{\dd z} \frac{T(z)^{k}}{k} \\
        &= \frac{n!}{k \cdot k!} [z^{n - 2}] \frac{\dd}{\dd z} T(z)^{k} \\
        &= \frac{n!}{k \cdot k!}(n - 1)[z^{n - 1}] T(z)^{k} \\
        &= \frac{n(n - 1)}{k} \cdot (n - 1)![z^{n - 1}] \frac{T(z)^{k}}{k!} \\
        &= \frac{n(n - 1)}{k} \binom{n - 2}{k - 1}(n - 1)^{n - 1 - k} \\
        &= (n - k)\binom{n}{k}(n - 1)^{n - 1 - k}
    \end{aligned}
\end{equation*}
This is the total number of degree-$k$ nodes over all rooted labelled trees of size $n$.
Since there are $n^{n-1}$ such trees, the mean is
\begin{equation*}
    \frac{(n - k)\binom{n}{k}(n - 1)^{n - 1 - k}}{n^{n-1}}
        \sim \frac{ne^{-1}}{k!}
\end{equation*}
for fixed $k$, matching the statement in Note III.17.

% Example III.15.
\exampleblock{III.15}{Path length in trees.}

For general Catalan trees, $G = \mathcal{Z}\times\mathrm{SEQ}(G)$, hence
\begin{equation*}
    G(z) = \frac{z}{1-G(z)}.
\end{equation*}
Let $\lambda(\tau)$ be the path length, the sum of distances of all nodes to the root. If
$\tau$ has root subtrees $\upsilon$, then
\begin{equation*}
    \lambda(\tau)=\sum_{\upsilon\prec\tau}\bigl(\lambda(\upsilon)+|\upsilon|\bigr).
\end{equation*}
The extra $|\upsilon|$ appears because every node in a root subtree is one level farther from
the root of $\tau$ than from the root of $\upsilon$.

Write $G(z,u)$ for the BGF where $u$ marks path length. Marking
$\lambda(\upsilon)+|\upsilon|$ instead of just $\lambda(\upsilon)$ corresponds to the substitution
$z\mapsto zu$, so the root decomposition gives
\begin{equation*}
    G(z,u)=\frac{z}{1-G(zu,u)}.
\end{equation*}
Now set
\begin{equation*}
    \Omega(z)=\left.\frac{\partial}{\partial u}G(z,u)\right|_{u=1}.
\end{equation*}
Differentiate the functional equation with respect to $u$:
\begin{equation*}
    G_u(z,u)
        =
        \frac{z}{(1-G(zu,u))^2}
        \left(zG_z(zu,u)+G_u(zu,u)\right).
\end{equation*}
Setting $u=1$ gives the linear equation
\begin{equation*}
    \Omega(z)=\frac{z}{(1-G(z))^2}\left(zG'(z)+\Omega(z)\right).
\end{equation*}
Solving for $\Omega(z)$ yields
\begin{equation*}
    \Omega(z)
        = \frac{z^2G'(z)}{(1-G(z))^2-z}.
\end{equation*}
It remains to simplify this expression:
\begin{equation*}
    z^{2} \frac{G'(z)}{(1 - G(z))^{2} - z} = \frac{z}{2(1 - 4z)} - \frac{z}{2\sqrt{1 - 4z}}
\end{equation*}
The clean route is to rewrite everything in terms of $G$:
\begin{equation*}
    G = \frac{z}{1 - G} \implies z = G(1 - G) = G - G^{2}
\end{equation*}
\begin{equation*}
    z = G - G^{2} \implies 1 = G' - 2GG' \implies G' = \frac{1}{1 - 2G}
\end{equation*}
Hence,
\begin{equation*}
    \begin{aligned}
        z^{2} \frac{G'(z)}{(1 - G(z))^{2} - z}
        &= (G(1 - G))^{2} \frac{1/(1 - 2G)}{1 - 2G + G^{2} - (G - G^{2})} \\
        &= \frac{G^{2}(1 - G)^{2}}{(1 - 2G)(1 - 3G + 2G^{2})} \\
        &= \frac{G^{2}(1 - G)}{(1 - 2G)^{2}} \\
        &= \frac{(G - G^{2}) \cdot G}{1 - 4(G - G^{2})} \\
        &= \frac{zG}{1 - 4z} \\
        &= \frac{z}{1 - 4z} \cdot \frac{1 - \sqrt{1 - 4z}}{2} \\
        &= \frac{z}{2} \left(\frac{1}{1 - 4z} - \frac{1}{\sqrt{1 - 4z}}\right)
    \end{aligned}
\end{equation*}

% Note III.20.
\noteblock{III.20}{Path length in simple varieties of trees.}

The same path-length recursion used for Catalan trees works for any simple variety:
when a subtree is attached below the root, all its nodes move one level down, so its contribution
changes from $\lambda$ to $\lambda + |\cdot|$. This is again the substitution $z\mapsto zu$
inside the recursive call. Thus we start from
\begin{equation*}
    T(z) = z\phi(T(z)) \; \text{where} \; T(z) = T(z, 1); \quad
    T(z, u) = z\phi(T(zu, u))
\end{equation*}
Let
\begin{equation*}
    \Omega(z) = T_u(z,1)
\end{equation*}
be the cumulative GF of path length.
Differentiate the second equation with respect to $u$:
\begin{equation*}
    \frac{\partial}{\partial u} T(z, u)
        = z\phi'(T(zu, u)) \cdot \frac{\dd}{\dd u} T(zu, u) = z\phi'(T(zu, u))[T_z(zu, u)z + T_u(zu, u)]
\end{equation*}
At $u = 1$, this becomes
\begin{equation} \label{eq:III.20.1}
    T_u(z, 1) = z\phi'(T(z))[zT'(z) + T_u(z, 1)] \implies T_u(z, 1) = \frac{z^{2}\phi'(T(z))T'(z)}{1 - z\phi'(T(z))}
\end{equation}
Differentiating the first equation with respect to $z$ gives:
\begin{equation*}
    T'(z) = \phi(T(z)) + z\phi'(T(z))T'(z) \implies \frac{1}{1 - z\phi'(T(z))} = \frac{T'(z)}{\phi(T(z))}
\end{equation*}
Insert this into \eqref{eq:III.20.1}:
\begin{equation*}
    \Omega(z) = T_u(z, 1) = z^{2}\phi'(T(z))T'(z) \cdot \frac{T'(z)}{\phi(T(z))} = \frac{\phi'(T(z))}{\phi(T(z))}(zT'(z))^{2}
\end{equation*}

\section{Complete generating functions and discrete models}
% Example III.16.
\exampleblock{III.16}{Words and records.}
After setting all letter markers to $1$, the cumulative GF for records is
\begin{equation*}
    \Omega(z)=\frac{z}{1-rz}\sum_{j=1}^{r}\frac{1}{1-(j-1)z}.
\end{equation*}
We only extract its coefficient and divide by the total number $r^n$ of words of length $n$.
\begin{equation*}
    \begin{aligned}
        [z^{n}] \frac{z}{1 - rz} \sum_{j = 1}^{r} \frac{1}{1 - (j - 1)z}
        &= [z^{n}] \frac{z}{1 - rz} \sum_{k = 0}^{r - 1} \frac{1}{1 - kz} \\
        &= [z^{n}] \sum_{k = 0}^{r - 1} \frac{z}{(1 - rz)(1 - kz)} \\
        &= [z^{n}] \sum_{k = 0}^{r - 1} \frac{1}{r - k} \left(\frac{rz}{1 - rz} - \frac{kz}{1 - kz} \right) \\
        &= \sum_{k = 0}^{r - 1} \frac{1}{r - k} \left([z^{n}] \frac{rz}{1 - rz} - [z^{n}] \frac{kz}{1 - kz} \right) \\
        &= \sum_{k = 0}^{r - 1} \frac{r^{n} - k^{n}}{r - k}
    \end{aligned}
\end{equation*}
\begin{equation*}
    \mathbb{E}_{\mathcal{W}_{n}}(\#\text{records})
    = \frac{1}{r^n}[z^n]\Omega(z)
    = \sum_{k = 0}^{r - 1} \frac{1 - (k/r)^{n}}{r - k}
    = \sum_{k = 0}^{r - 1} \frac{1}{r - k} - \sum_{k = 0}^{r - 1} \frac{(k/r)^{n}}{r - k}
    = H_{r} - \sum_{k = 1}^{r - 1} \frac{(k/r)^{n}}{r - k}
\end{equation*}

% Example III.19.
\exampleblock{III.19}{Rises in Bernoulli trials: Simon Newcomb's problem.}

Let $z_j=zu_j$, where $u_j$ marks occurrences of the letter $a_j$. Let $v$ mark weak rises:
a position followed by a letter that is at least as large as the current one. Write $W_j$ for
non-empty words that start with $a_j$, and write
\begin{equation*}
    W = 1 + W_1+\cdots+W_r
\end{equation*}
for all words, including the empty word.

\stephead{1}{Pass from $W_j$ to $X_j$.}
Start from the recurrence:
\begin{equation} \label{eq:III.19.1}
    W_{j} = z_{j} + z_{j} \sum_{k = 1}^{j - 1} W_{k} + vz_{j} \sum_{k = j}^{r} W_{k},
\end{equation}
The first term is the one-letter word $a_j$. If the next letter is smaller than $a_j$, there is
no weak rise, giving the middle sum. If the next letter is at least $a_j$, we get one weak rise,
giving the factor $v$ in the last sum.
It is convenient to pass to $X_j$:
\begin{equation} \label{eq:III.19.2}
    X_{j} = 1 + \sum_{k = 1}^{j - 1} W_{k} + v \sum_{k = j}^{r} W_{k} = 1 + \sum_{k = 1}^{j - 1} z_{k}X_{k} + v \sum_{k = j}^{r} z_{k}X_{k}.
\end{equation}
Now take the forward difference:
\begin{equation*}
    X_{j + 1} - X_{j} = W_{j} - vW_{j} = z_{j}X_{j}(1 - v),
\end{equation*}
which gives the first-order recurrence:
\begin{equation} \label{eq:III.19.4}
    X_{j + 1} = X_{j}(1 + z_{j}(1 - v)).
\end{equation}

\stephead{2}{Solve the first-order recurrence.}
Unwinding \eqref{eq:III.19.4} shows that every $X_j$ is determined by $X_1$:
\begin{equation} \label{eq:III.19.3}
    X_{j} = X_{1} \prod_{k = 1}^{j - 1}(1 + (1 - v)z_{k}).
\end{equation}

\stephead{3}{Determine $X_1$.}
To determine $X_1$, look at the boundary case $j = 1$:
\begin{equation*}
    W_{1} = z_{1} + vz_{1} \sum_{k = 1}^{r} W_{k} \implies z_{1}X_{1} = z_{1} + vz_{1} \sum_{k = 1}^{r} z_{k}X_{k}.
\end{equation*}
Canceling $z_{1}$ gives:
\begin{equation*}
    X_{1} = 1 + v \sum_{k = 1}^{r} z_{k}X_{k}.
\end{equation*}
To evaluate the remaining sum, introduce the formal boundary term $X_{r + 1}$ by setting $j = r + 1$ in \eqref{eq:III.19.2}. Then the second sum disappears, leaving
\begin{equation*}
    X_{r + 1} = 1 + \sum_{k = 1}^{r} z_{k}X_{k} \implies \sum_{k = 1}^{r} z_{k}X_{k} = X_{r + 1} - 1.
\end{equation*}
Substituting back gives:
\begin{equation*}
    X_{1} = 1 + v(X_{r + 1} - 1).
\end{equation*}
On the other hand, \eqref{eq:III.19.3} at $j = r + 1$ gives $X_{r + 1} = X_{1}P$, where
$P = \prod_{j = 1}^{r}(1 + (1 - v)z_{j})$. Plugging this into the equation for $X_1$,
\begin{equation*}
    X_{1} = 1 + v(X_{1}P - 1),
\end{equation*}
and solving for $X_1$ gives:
\begin{equation} \label{eq:III.19.5}
    X_{1} = \frac{1 - v}{1 - vP}.
\end{equation}

\stephead{4}{Recover the total generating function.}
Finally, the total generating function is:
\begin{equation} \label{eq:III.19.6}
    W = 1 + \sum_{k = 1}^{r} W_{k} = 1 + \sum_{k = 1}^{r} z_{k}X_{k} = X_{r + 1} = \frac{P(1 - v)}{1 - vP}.
\end{equation}
This is the same as the book's form
\begin{equation*}
    W = \frac{v - 1}{v - P^{-1}},
\end{equation*}
after multiplying numerator and denominator by $P$.

\section{Additional constructions}
% Example III.22.
\exampleblock{III.22}{Constrained integer compositions and “slicing”.}

This example concerns two independent regions, $\mathcal{R}_{1}$ and $\mathcal{R}_{2}$. They illustrate the same slicing philosophy, but the functional equations and resulting series are derived separately.

\subhead{Region $\mathcal{R}_{1}$}
\stephead{1}{Derive the functional equation for $\mathcal{R}_{1}$.}
\begin{equation*}
    \mathcal{R}_{1} = \{(x, y)|1 \leq x \leq y\}
\end{equation*}
To obtain the functional equation, note that
\begin{equation*}
    \begin{aligned}
        f(z) &= z + z^{2} + \cdots = \frac{z}{1 - z} \\
        \mathcal{L}[F(z, u)]
            &= \mathcal{L}[\sum_{j} f^{\langle j \rangle}(z)u^{j}] \\
            &= \sum_{j} f^{\langle j \rangle}(z)\mathcal{L}[u^{j}] \\
            &= \sum_{j} f^{\langle j \rangle}(z)\frac{u^{j}}{1 - u} \\
            &= \frac{1}{1 - u} F(z, u)
    \end{aligned}
\end{equation*}
Therefore,
\begin{equation} \label{eq:III.22.3}
    F(z, u) = f(zu) + (\mathcal{L}[F(z, u)])_{u \mapsto zu} = \frac{zu}{1 - zu} + \frac{1}{1 - zu} F(z, zu)
\end{equation}

\stephead{2}{Check the closed form for $\mathcal{R}_{1}$.}
Now let us check that the two advertised forms of $F(z, u)$ really agree:
\begin{equation} \label{eq:III.22.1}
    F(z, u) = \frac{zu}{1 - zu} + \frac{z^{2}u}{(1 - zu)(1 - z^{2}u)} + \frac{z^{3}u}{(1 - zu)(1 - z^{2}u)(1 - z^{3}u)} + \cdots
\end{equation}
\begin{equation} \label{eq:III.22.2}
    F(z, u) = \frac{zu}{1 - z} + \frac{z^{2}u^{2}}{(1 - z)(1 - z^{2})} + \frac{z^{3}u^{3}}{(1 - z)(1 - z^{2})(1 - z^{3})} + \cdots
\end{equation}
\eqref{eq:III.22.1} comes from repeated substitution. One could verify \eqref{eq:III.22.2} by series expansion, but there is a cleaner algebraic route.
Assume $F(z, u) = \sum_{k \geq 1} A_{k}(z)u^{k}$ and substitute into the functional equation \eqref{eq:III.22.3}:
\begin{equation*}
    (1 - zu)F(z, u) = zu + F(z, zu)
\end{equation*}
Expanding both sides gives:
\begin{equation*}
    \sum_{k \geq 1} A_{k}(z)u^{k} - zu\left(\sum_{k \geq 1} A_{k}(z)u^{k}\right) = zu + \sum_{k \geq 1} A_{k}(z)(zu)^{k}
\end{equation*}
Now collect powers of $u$:
\begin{equation*}
    \sum_{k \geq 1} A_{k}(z)u^{k} - \sum_{k \geq 1} A_{k}(z)zu^{k + 1} = zu + \sum_{k \geq 1} A_{k}(z)z^{k}u^{k}
\end{equation*}
Extract the coefficients of $u^{k}$ (for $k > 1$):
\begin{equation} \label{eq:III.22.4}
    A_{k}(z) - zA_{k - 1}(z) = z^{k}A_{k}(z) \implies A_{k}(z) = \frac{z}{1 - z^{k}}A_{k - 1}(z)
\end{equation}
Starting from $A_{1}(z) = z + z^{2} + \cdots = \frac{z}{1 - z}$, the recurrence telescopes to
\begin{equation*}
    A_{k}(z) = \frac{z}{1 - z^{k}} \frac{z}{1 - z^{k - 1}} \cdots \frac{z}{1 - z^{2}} \frac{z}{1 - z}
        = \frac{z^{k}}{(1 - z)(1 - z^{2}) \cdots (1 -z^{k})}
\end{equation*}
So the $k$th term of \eqref{eq:III.22.2} is exactly $A_{k}(z)u^{k}$.

\stephead{3}{Compare the two combinatorial interpretations.}
There is also a nice combinatorial punchline: the $k$th term of \eqref{eq:III.22.1} counts partitions with exactly $k$ summands, whereas the $k$th term of \eqref{eq:III.22.2} counts partitions whose largest summand is exactly $k$.
The case $k = 3$ makes the contrast visible.

The 3rd term in \eqref{eq:III.22.1}:
\begin{equation*}
    \frac{z^{3}u}{(1 - zu)(1 - z^{2}u)(1 - z^{3}u)} = zz(zu)\sum_{j}(zu)^{j} \sum_{k} z^{k}(zu)^{k} \sum_{l} z^{l}z^{l}(zu)^{l}
\end{equation*}
\begin{itemize}
    \item $z^{1} \cdot z^{1} \cdot (zu)^{1}$ places $1$ cell on all three columns
    \item $z^{0} \cdot z^{0} \cdot (zu)^{j}$ places $j \geq 0$ cells on the rightmost column
    \item $z^{0} \cdot z^{k} \cdot (zu)^{k}$ places $k \geq 0$ cells on the rightmost two columns
    \item $z^{l} \cdot z^{l} \cdot (zu)^{l}$ places $l \geq 0$ cells on all three columns
\end{itemize}
This lifting mechanism is what keeps the summands weakly increasing.

The 3rd term in \eqref{eq:III.22.2}:
\begin{equation*}
    \frac{z^{3}u^{3}}{(1 - z)(1 - z^{2})(1 - z^{3})}
\end{equation*}
\begin{itemize}
    \item $z^{3}u^{3}$ ensures there is at least one part of size $3$ (which also marks the last column height $3$).
    \item $\frac{1}{(1 - z)(1 - z^{2})(1 - z^{3})}$ is the standard generating function for partitions using parts of size $\{1, 2, 3\}$.
\end{itemize}

We now leave $\mathcal{R}_{1}$ entirely and start a separate analysis for $\mathcal{R}_{2}$.

\subhead{Region $\mathcal{R}_{2}$}
\stephead{1}{Derive the functional equation for $\mathcal{R}_{2}$.}
\begin{equation*}
    \mathcal{R}_{2} = \{(x, y)|1 \leq y \leq 2x\}
\end{equation*}
For this second region, the slicing operator changes, so we begin again:
\begin{equation*}
    \begin{aligned}
        f(z) &= z \\
        \mathcal{L}[F(z, u)]
            &= \sum_{j} f^{\langle j \rangle}(z)\mathcal{L}[u^{j}] \\
            &= \sum_{j} f^{\langle j \rangle}(z)u \frac{1 - u^{2j}}{1 - u} \\
            &= \frac{u}{1 - u} \sum_{j} f^{\langle j \rangle}(z)(1 - u^{2j}) \\
            &= \frac{u}{1 - u} \left(\sum_{j} f^{\langle j \rangle}(z) - \sum_{j} f^{\langle j \rangle}(z)(u^{2})^{j}\right) \\
            &= \frac{u}{1 - u} \left(F(z, 1) - F(z, u^{2})\right)
    \end{aligned}
\end{equation*}
Hence,
\begin{equation} \label{eq:III.22.5}
    F(z, u) = f(zu) + (\mathcal{L}[F(z, u)])_{u \mapsto zu} = zu + \frac{zu}{1 - zu} \left(F(z, 1) - F(z, z^{2}u^{2})\right)
\end{equation}
Repeated substitution gives:
\begin{equation*}
    F(z, u) = \alpha(u) - \frac{zu}{1 - zu} \alpha(z^{2}u^{2}) + \frac{zu}{1 - zu} \frac{z^{3}u^{2}}{1 - z^{3}u^{2}}\alpha(z^{6}u^{4}) - \cdots
\end{equation*}
where
\begin{equation*}
    \alpha(u) = zu + \frac{zu}{1 - zu} F(z, 1), \quad \beta(u) = -\frac{zu}{1 - zu}, \quad \sigma(u) = z^{2}u^{2}
\end{equation*}

\stephead{2}{Solve for $F(z, 1)$.}
Set $u = 1$ and write $X = F(z, 1)$:
\begin{equation*}
    \begin{aligned}
        X
        &= \alpha(1) - \frac{z}{1 - z} \alpha(z^{2}) + \frac{z}{1 - z} \frac{z^{3}}{1 - z^{3}}\alpha(z^{6}) - \cdots \\
        &= \left[z + \frac{z}{1 - z} X\right]
            - \frac{z}{1 - z} \left[z^{3} + \frac{z^{3}}{1 - z^{3}}X\right]
            + \frac{z}{1 - z} \frac{z^{3}}{1 - z^{3}}\left[z^{7} + \frac{z^{7}}{1 - z^{7}}X\right] - \cdots
    \end{aligned}
\end{equation*}
This is a linear equation in $X$:
\begin{equation} \label{eq:III.22.6}
    X = A(z) + X B(z).
\end{equation}
From here, separate the constant contribution from the coefficient of $X$. This gives
\begin{equation*}
    \begin{aligned}
        A(z)
        &= z - \frac{z}{1 - z} z^{3} + \frac{z}{1 - z} \frac{z^{3}}{1 - z^{3}}z^{7} - \cdots \\
        &= z - \frac{z^{1 + 3}}{1 - z} + \frac{z^{1 + 3 + 7}}{(1 - z)(1 - z^{3})} - \cdots \\
        &= \sum_{j \geq 1}(-1)^{j - 1}z^{2^{j + 1} - j - 2}/\left((1 - z)(1 - z^{3}) \cdots (1 - z^{2^{j - 1} - 1})\right) \\
        &= \sum_{j \geq 1}(-1)^{j - 1}z^{2^{j + 1} - j - 2}/Q_{j - 1}(z)
    \end{aligned}
\end{equation*}
The exponent is:
\begin{equation*}
    \begin{aligned}
        1 + 3 + 7 + \cdots + (2^{j} - 1)
        &= (2 - 1) + (2^{2} - 1) + (2^{3} - 1) + \cdots + (2^{j} - 1) \\
        &= 2 + 2^{2} + 2^{3} + \cdots + 2^{j} - j \\
        &= \frac{2 - 2^{j + 1}}{1 - 2} - j \\
        &= 2^{j + 1} - j - 2
    \end{aligned}
\end{equation*}
\begin{equation*}
    \begin{aligned}
        1 - B(z)
        &= 1 - \frac{z}{1 - z} + \frac{z}{1 - z} \frac{z^{3}}{1 - z^{3}} - \frac{z}{1 - z} \frac{z^{3}}{1 - z^{3}} \frac{z^{7}}{1 - z^{7}} + \cdots \\
        &= 1 - \frac{z}{1 - z} + \frac{z^{1 + 3}}{(1 - z)(1 - z^{3})} - \frac{z^{1 + 3 + 7}}{(1 - z)(1 - z^{3})(1 - z^{7})} + \cdots \\
        &= \sum_{j \geq 0}(-1)^{j}z^{2^{j + 1} - j - 2}/\left((1 - z)(1 - z^{3}) \cdots (1 - z^{2^{j} - 1})\right) \\
        &= \sum_{j \geq 0}(-1)^{j}z^{2^{j + 1} - j - 2}/Q_{j}(z)
    \end{aligned}
\end{equation*}
Finally,
\begin{equation} \label{eq:III.22.7}
    F(z, 1) = X = \frac{\sum_{j \geq 1}(-1)^{j - 1}z^{2^{j + 1} - j - 2}/Q_{j - 1}(z)}{\sum_{j \geq 0}(-1)^{j}z^{2^{j + 1} - j - 2}/Q_{j}(z)}
\end{equation}
where
\begin{equation*}
    Q_j(z) = \prod_{m = 1}^{j}\left(1 - z^{2^m - 1}\right)
        = (1 - z)(1 - z^3)(1 - z^7)\cdots(1 - z^{2^j - 1}),
        \qquad Q_0(z)=1.
\end{equation*}

This conclusion belongs only to $\mathcal{R}_{2}$; it does not depend on the earlier computation for $\mathcal{R}_{1}$.

The sequence begins
$1,1,2,3,5,9,16,28,50$ and is OEIS A002572. One interpretation is:
the coefficient indexed by $m$ counts partitions of $1$ into exactly $m$ summands chosen from
\[
    \frac{1}{2},\frac{1}{4},\frac{1}{8},\ldots
\]
Here the summand $1$ should not be included; otherwise the first terms would be shifted by
the trivial partition $1=1$. For example, there is $1$ way with $2$ summands,
$1$ way with $3$ summands, $2$ ways with $4$ summands,
$3$ ways with $5$ summands, and $5$ ways with $6$ summands:
\begin{equation*}
    \begin{aligned}
        1 &= \frac{1}{2} + \frac{1}{2} \\[20pt]
        1 &= \frac{1}{2} + \frac{1}{4} + \frac{1}{4} \\[20pt]
        1 &= \frac{1}{4} + \frac{1}{4} + \frac{1}{4} + \frac{1}{4} \\
        1 &= \frac{1}{2} + \frac{1}{4} + \frac{1}{8} + \frac{1}{8} \\[20pt]
        1 &= \frac{1}{4} + \frac{1}{4} + \frac{1}{4} + \frac{1}{8} + \frac{1}{8} \\
        1 &= \frac{1}{2} + \frac{1}{8} + \frac{1}{8} + \frac{1}{8} + \frac{1}{8} \\
        1 &= \frac{1}{2} + \frac{1}{4} + \frac{1}{8} + \frac{1}{16} + \frac{1}{16} \\[20pt]
        1 &= \frac{1}{4} + \frac{1}{4} + \frac{1}{8} + \frac{1}{8} + \frac{1}{8} + \frac{1}{8} \\
        1 &= \frac{1}{4} + \frac{1}{4} + \frac{1}{4} + \frac{1}{8} + \frac{1}{16} + \frac{1}{16} \\
        1 &= \frac{1}{2} + \frac{1}{8} + \frac{1}{8} + \frac{1}{8} + \frac{1}{16} + \frac{1}{16} \\
        1 &= \frac{1}{2} + \frac{1}{4} + \frac{1}{16} + \frac{1}{16} + \frac{1}{16} + \frac{1}{16} \\
        1 &= \frac{1}{2} + \frac{1}{4} + \frac{1}{8} + \frac{1}{16} + \frac{1}{32} + \frac{1}{32}
    \end{aligned}
\end{equation*}

% Note III.32.
\noteblock{III.32}{Carlitz compositions I.}

To obtain the functional equation, note that
\begin{equation*}
    \begin{aligned}
        k(z) &= z + z^{2} + \cdots = \frac{z}{1 - z} \\
        \mathcal{L}[K(z, u)]
            &= \sum_{j} k^{\langle j \rangle}(z)\mathcal{L}[u^{j}] \\
            &= \sum_{j} k^{\langle j \rangle}(z)\left(\frac{u}{1 - u} - u^{j}\right) \\
            &= \frac{u}{1 - u} \sum_{j} k^{\langle j \rangle}(z) - \sum_{j} k^{\langle j \rangle}(z)u^{j} \\
            &= \frac{u}{1 - u} K(z, 1) - K(z, u)
    \end{aligned}
\end{equation*}
Therefore,
\begin{equation*}
    K(z, u) = k(zu) + (\mathcal{L}[K(z, u)])_{u \mapsto zu} = \frac{zu}{1 - zu} + \frac{zu}{1 - zu} K(z, 1) - K(z, zu)
\end{equation*}
To solve for $K(z, 1)$, write $X = K(z, 1)$:
\begin{equation*}
    K(z, u) = \frac{zu}{1 - zu}(1 + X) - K(z, zu)
\end{equation*}
One iteration gives:
\begin{equation*}
    K(z, u) = \frac{zu}{1 - zu}(1 + X) - \left(\frac{z^{2}u}{1 - z^{2}u}(1 + X) - K(z, z^{2}u)\right)
\end{equation*}
Iterating all the way gives:
\begin{equation*}
    \begin{aligned}
        K(z, u)
        &= \frac{zu}{1 - zu}(1 + X) - \frac{z^{2}u}{1 - z^{2}u}(1 + X) + \frac{z^{3}u}{1 - z^{3}u}(1 + X) - \cdots \\
        &= (1 + X)\sum_{j \geq 1}(-1)^{j - 1} \frac{z^{j}u}{1 - z^{j}u}
    \end{aligned}
\end{equation*}
Now set $u = 1$:
\begin{equation*}
    X = (1 + X)\sum_{j \geq 1} \frac{(-1)^{j - 1}z^{j}}{1 - z^{j}}
\end{equation*}
If we call the sum $S$, then
\begin{equation*}
    X = (1 + X)S \implies 1 + X = \frac{1}{1 - S} = \left(1 - \sum_{j \geq 1} \frac{(-1)^{j - 1}z^{j}}{1 - z^{j}}\right)^{-1}
        = \left(1 + \sum_{j \geq 1} \frac{(-z)^{j}}{1 - z^{j}}\right)^{-1}
\end{equation*}
The small bookkeeping point is that this final expression equals $1 + X$, not $X$ itself.
In other words, it is the GF where the empty composition is included. For the non-empty class
tracked by $K(z,u)$, the answer is instead
\begin{equation*}
    K(z,1)=X
        = \left(1 + \sum_{j \geq 1} \frac{(-z)^j}{1-z^j}\right)^{-1} - 1.
\end{equation*}
This also explains the constant term in the book's displayed series
\[
    1 + z + z^2 + 3z^3 + 4z^4 + \cdots:
\]
that series is $1+K(z,1)$.

% Example III.23.
\exampleblock{III.23}{Local order patterns in permutations.}

\stephead{1}{Solve the differential equation.}
For non-empty increasing binary trees, the boxed-product specification gives
\begin{equation*}
    \frac{\dd}{\dd z}\hat{I}(z,\mathbf{u})
        = u_0 + (u_1+u_1')\hat{I}(z,\mathbf{u}) + u_2\hat{I}(z,\mathbf{u})^2,
        \qquad \hat{I}(0,\mathbf{u})=0.
\end{equation*}
For brevity, let $\hat{I}(z, \mathbf{u})$ be abbreviated to $I$, and set
\begin{equation*}
    v_1=\frac{u_1+u_1'}{2}, \qquad \delta^2=u_0u_2-v_1^2.
\end{equation*}
Then the differential equation becomes
\begin{equation*}
    \frac{\dd I}{\dd z} = u_{0} + (u_{1} + u_{1}')I + u_{2}I^{2} = u_{0} + 2v_{1}I + u_{2}I^{2}
\end{equation*}
Separating variables yields:
\begin{equation*}
    \frac{\dd I}{u_{0} + 2v_{1}I + u_{2}I^{2}} = \dd z
\end{equation*}
Integrate this from the initial condition $I(0) = 0$ up to a general $z$:
\begin{equation*}
    \int_{\xi = I(0)}^{\xi = I(z)} \frac{\dd \xi}{u_{0} + 2v_{1}\xi + u_{2}\xi^{2}} = \int_{\zeta = 0}^{\zeta = z} \dd \zeta
\end{equation*}
Using $I(0) = 0$ and writing $I(z)$ simply as $I$,
\begin{equation*}
    \int_{0}^{I} \frac{\dd \xi}{u_{0} + 2v_{1}\xi + u_{2}\xi^{2}} = \int_{0}^{z} \dd \zeta = z
\end{equation*}
Complete the square in the denominator:
\begin{equation*}
    u_{0} + 2v_{1}\xi + u_{2}\xi^{2}
    = u_{2}\left[\left(\xi + \frac{v_{1}}{u_{2}}\right)^{2} + \frac{\delta^{2}}{u_{2}^{2}}\right]
\end{equation*}
which turns the integral into
\begin{equation*}
    \int_{0}^{I} \frac{\dd \xi}{u_{2}\left[\left(\xi + \frac{v_{1}}{u_{2}}\right)^{2} + \left(\frac{\delta}{u_{2}}\right)^{2}\right]} = z
\end{equation*}
Using the standard integral $\int \frac{\dd x}{x^{2} +a^{2}} = \frac{1}{a} \arctan \frac{x}{a}$:
\begin{equation*}
    z = \frac{1}{u_{2}} \frac{1}{\frac{\delta}{u_{2}}}
        \arctan\left.\frac{\xi + \frac{v_{1}}{u_{2}}}{\frac{\delta}{u_{2}}}\right|_{\xi = 0}^{\xi = I}
      = \frac{1}{\delta} \left[
            \arctan\left(\frac{u_{2}I + v_{1}}{\delta}\right) - \arctan\left(\frac{v_{1}}{\delta}\right)
        \right]
\end{equation*}
Now rearrange and take tangents:
\begin{equation*}
    \frac{u_{2}I + v_{1}}{\delta} = \tan\left[\arctan\left(\frac{v_{1}}{\delta}\right) + z\delta\right]
        = \frac{\frac{v_{1}}{\delta} + \tan(z\delta)}{1 - \frac{v_{1}}{\delta}\tan(z\delta)}
        = \frac{v_{1} + \delta\tan(z\delta)}{\delta - v_{1}\tan(z\delta)}
\end{equation*}
Solving for $I$ yields:
\begin{equation} \label{eq:III.23.2}
    I = \frac{\delta}{u_{2}} \frac{v_{1} + \delta\tan(z\delta)}{\delta - v_{1}\tan(z\delta)} - \frac{v_{1}}{u_{2}}
\end{equation}

\stephead{2}{Expand the closed form.}
Now expand $I$ as a power series. Start from the closed form \eqref{eq:III.23.2}:

The following algebra temporarily divides by $v_1$; the final expansion is a polynomial identity,
so this is just a convenient way to organize the terms.
\begin{equation*}
    \begin{aligned}
        I
        &= \frac{v_{1}}{u_{2}} \frac{1 + \frac{\delta}{v_{1}}\tan(z\delta)}{1 - \frac{v_{1}}{\delta}\tan(z\delta)} - \frac{v_{1}}{u_{2}} \\
        &= \frac{v_{1}}{u_{2}}\left[1 + \frac{\delta}{v_{1}}\tan(z\delta)\right]
            \left[1 + \frac{v_{1}}{\delta}\tan(z\delta)
                + \left(\frac{v_{1}}{\delta}\tan(z\delta)\right)^{2}
                + \left(\frac{v_{1}}{\delta}\tan(z\delta)\right)^{3} + \cdots \right]
             - \frac{v_{1}}{u_{2}} \\
        &= \frac{v_{1}}{u_{2}}\left[1 + \frac{\delta}{v_{1}}\left(z\delta + \frac{(z\delta)^{3}}{3} + \cdots \right)\right]
            \left[1 + \frac{v_{1}}{\delta}\left(z\delta + \frac{(z\delta)^{3}}{3}\right)
                + \left(\frac{v_{1}}{\delta}(z\delta)\right)^{2}
                + \left(\frac{v_{1}}{\delta}(z\delta)\right)^{3} + \cdots \right] - \frac{v_{1}}{u_{2}} \\
        &= \frac{v_{1}}{u_{2}}\left[1 + \frac{\delta^{2}}{v_{1}}z + \frac{\delta^{4}}{3v_{1}}z^{3} + \cdots \right]
            \left[1 + v_{1}z + v_{1}^{2}z^{2} + \left(\frac{v_{1}\delta^{2}}{3} + v_{1}^{3}\right)z^{3} + \cdots \right] - \frac{v_{1}}{u_{2}} \\
        &= u_{0}z + u_{0}(u_{1} + u_{1}')\frac{z^{2}}{2!} + u_{0}((u_{1} + u_{1}')^{2} + 2u_{0}u_{2})\frac{z^{3}}{3!} + \cdots
    \end{aligned}
\end{equation*}
The same coefficients could also be recovered directly from the differential equation by repeated differentiation, but the closed form is shorter.

\stephead{3}{Check the first few objects directly.}
There is also a direct check: list the objects for small $n$.
\begin{equation*}
    \begin{aligned}
        \mathcal{P}_{1} &= \{ \cnum{1} \} \implies \\
        f_{1}(\mathbf{u}) &= u_{0} \\[10pt]
        \mathcal{P}_{2} &= \{ \cnum{1}\text{-}\cnum{2}, \cnum{2}\text{-}\cnum{1} \} \implies \\
        f_{2}(\mathbf{u}) &= u_{1}u_{0} + u_{0}u_{1}' = u_{0}(u_{1} + u_{1}') \\[10pt]
        \mathcal{P}_{3} &= \{ \cnum{1}\text{-}\cnum{2}\text{-}\cnum{3}, \cnum{1}\text{-}\cnum{3}\text{-}\cnum{2},
            \cnum{2}\text{-}\cnum{1}\text{-}\cnum{3}, \cnum{2}\text{-}\cnum{3}\text{-}\cnum{1},
            \cnum{3}\text{-}\cnum{1}\text{-}\cnum{2}, \cnum{3}\text{-}\cnum{2}\text{-}\cnum{1} \} \implies \\
        f_{3}(\mathbf{u}) &= u_{1}u_{1}u_{0} + u_{1}u_{0}u_{1}' + u_{0}u_{2}u_{0} + u_{1}u_{0}u_{1}' + u_{0}u_{2}u_{0} + u_{0}u_{1}'u_{1}' \\
            &= u_{0}(u_{1}^{2} + u_{1}u_{1}' + u_{0}u_{2} + u_{1}u_{1}' + u_{0}u_{2} + u_{1}'^{2}) \\
            &= u_{0}((u_{1} + u_{1}')^{2} + 2u_{0}u_{2}) \\[10pt]
        \mathcal{P}_{4} &= \{ \cdots \} \\
        f_{4}(\mathbf{u}) &= u_{0}(u_{1} + u_{1}')((u_{1} + u_{1}')^{2} + 8u_{0}u_{2})
    \end{aligned}
\end{equation*}
Indeed, if $a=u_1+u_1'$ and $\hat{I}(z,\mathbf{u})=\sum f_n(\mathbf{u})z^n/n!$, the
differential equation gives
\[
    f_4 = af_3 + u_2\left(3f_1f_2+3f_2f_1\right)
        = au_0(a^2+2u_0u_2)+6au_0^2u_2
        = au_0(a^2+8u_0u_2).
\]

With the closed form established, we can now extract the mean numbers of the three node types by specializing one marker at a time.

\subhead{Nullary Type}
Set $u_{1} = u_{1}' = u_{2} = 1$. Then $v_{1} = 1$ and $\delta = \sqrt{u_{0} - 1}$. We have
\begin{equation*}
    \hat{I}(z, u_{0}) = \delta \frac{1 + \delta\tan(z\delta)}{\delta - \tan(z\delta)} - 1 = (1 + \delta^{2})\frac{w}{1 - w}
\end{equation*}
where $w \coloneqq \frac{\tan(z\delta)}{\delta} = z + \frac{z^{3}}{3}\delta^{2} + O(\delta^{4})$.
Write $w = w_{0} + w_{2}\delta^{2} + O(\delta^{4})$ with $w_{0} = z$ and $w_{2} = z^{3}/3$. Then
\begin{equation*}
    1 - w = 1 - w_{0} - w_{2}\delta^{2} + O(\delta^{4}) = (1 - w_{0})\left[1 - \frac{w_{2}}{1 - w_{0}}\delta^{2} + O(\delta^{4})\right]
\end{equation*}
Hence,
\begin{equation*}
    \frac{1}{1 - w} = \frac{1}{1 - w_{0}} \frac{1}{1 - \frac{w_{2}}{1 - w_{0}}\delta^{2} + \cdots}
        = \frac{1}{1 - w_{0}}\left[1 + \frac{w_{2}}{1 - w_{0}}\delta^{2} + O(\delta^{4})\right]
\end{equation*}
Multiplying by $w = w_{0} + w_{2}\delta^{2} + O(\delta^{4})$ yields:
\begin{equation*}
    \frac{w}{1 - w} = \frac{w_{0}}{1 - w_{0}} + \frac{w_{2}}{(1 - w_{0})^{2}}\delta^{2} + O(\delta^{4})
\end{equation*}
Substituting $w_{0} = z$ and $w_{2} = z^{3}/3$, we obtain
\begin{equation} \label{eq:III.23.1}
    \frac{w}{1 - w} = \frac{z}{1 - z} + \frac{z^{3}}{3(1 - z)^{2}}\delta^{2} + O(\delta^{4})
\end{equation}
Since $\delta^{2} = u_{0} - 1$, the first-order Taylor term in $u_{0}$ is exactly the coefficient of $\delta^{2}$. Hence,
\begin{equation*}
    \left.\frac{\partial I}{\partial u_{0}}\right|_{u_{0} = 1}
        = [\delta^{2}] \left((1 + \delta^{2})\frac{w}{1 - w} \right)
        = \frac{z}{1 - z} + \frac{z^{3}}{3(1 - z)^{2}}
\end{equation*}
For $n \geq 3$, the coefficient of $z^{n}$ in
$\left.\frac{\partial I}{\partial u_{0}}\right|_{u_{0} = 1}$ is $(n + 1)/3$.

\subhead{Unary Type}
Treat the two unary types together. Note that $\hat{I}$ depends on $u_{1}, u_{1}'$ only through
$v_{1} = (u_{1} + u_{1}')/2$. Set $u_{0} = u_{2} = 1$. Then $\delta = \sqrt{1 - v_{1}^{2}}$ and
\begin{equation*}
    \hat{I}(z, v_{1}) = \delta \frac{v_{1} + \delta\tan(z\delta)}{\delta - v_{1}\tan(z\delta)} - v_{1}
        = \frac{v_{1} + \delta^{2}w}{1 - v_{1}w} - v_{1} = \frac{w}{1 - v_{1}w}
\end{equation*}
where $w \coloneqq \frac{\tan(z\delta)}{\delta} = z + \frac{z^{3}}{3}\delta^{2} + O(\delta^{4})$.
We have
\begin{equation*}
    \frac{\partial I}{\partial u_{1}} = \frac{\partial I}{\partial u_{1}'} = \frac{1}{2} \frac{\partial I}{\partial v_{1}}
\end{equation*}
So it is enough to compute $\frac{\partial I}{\partial v_{1}}$ at $v_{1} = 1$. First, $w$ depends on $v_{1}$ through $s \coloneqq \delta^{2} = 1 - v_{1}^{2}$:
\begin{equation*}
    \frac{\partial w}{\partial v_{1}} = \frac{\partial w}{\partial s} \cdot (-2v_{1})
\end{equation*}
From the series $\left.\frac{\partial w}{\partial s} \right|_{s = 0} = \frac{z^{3}}{3}$, so at $v_{1} = 1$:
\begin{equation*}
    \left.\frac{\partial w}{\partial v_{1}}\right|_{v_{1} = 1} = -\frac{2z^{3}}{3}
\end{equation*}
Now differentiate $\hat{I}(z, v_{1})$:
\begin{equation*}
    \frac{\partial \hat{I}}{\partial v_{1}}
        = \frac{\frac{\partial w}{\partial v_{1}}(1 - v_{1}w) + w(w + v_{1} \frac{\partial w}{\partial v_{1}})}{(1 - v_{1}w)^{2}}
\end{equation*}
Substituting $v_{1} = 1, w = z$ and $\frac{\partial w}{\partial v_{1}} = -\frac{2z^{3}}{3}$, we obtain
\begin{equation*}
    \left.\frac{\partial \hat{I}}{\partial v_{1}}\right|_{v_{1} = 1} = \frac{z^{2}\left(1 - \frac{2}{3} z\right)}{(1 - z)^{2}}
\end{equation*}
Consequently
\begin{equation*}
    \left.\frac{\partial \hat{I}}{\partial u_{1}}\right|_{u_{1} = 1} = \left.\frac{\partial \hat{I}}{\partial u_{1}'}\right|_{u_{1}' = 1}
        = \frac{z^{2}\left(1 - \frac{2}{3} z\right)}{2(1 - z)^{2}}
\end{equation*}
For $n \geq 3$, the coefficient of $z^{n}$ in
$\left.\frac{\partial \hat{I}}{\partial u_{1}}\right|_{u_{1} = 1}$ and
$\left.\frac{\partial \hat{I}}{\partial u_{1}'}\right|_{u_{1}' = 1}$ is $(n + 1)/6$.

\subhead{Binary Type}
Set $u_{0} = 1$, $v_{1} = (1 + 1)/2 = 1$. Then $\delta = \sqrt{u_{2} - 1}$ and
\begin{equation*}
    \hat{I}(z, u_{2}) = \frac{\delta}{u_{2}} \frac{1 + \delta\tan(z\delta)}{\delta - \tan(z\delta)} - \frac{1}{u_{2}} = \frac{w}{1 - w}
\end{equation*}
where $w \coloneqq \frac{\tan(z\delta)}{\delta} = z + \frac{z^{3}}{3}\delta^{2} + O(\delta^{4})$.
This is the same calculation as in the nullary case, with one parameter renamed. From \eqref{eq:III.23.1} we obtain
\begin{equation*}
    \left.\frac{\partial \hat{I}}{\partial u_{2}}\right|_{u_{2} = 1}
    = [\delta^{2}] \frac{w}{1 - w}
    = \frac{z^{3}}{3(1 - z)^{2}}
\end{equation*}
For $n \geq 3$, the coefficient of $z^{n}$ in
$\left.\frac{\partial \hat{I}}{\partial u_{2}}\right|_{u_{2} = 1}$ is $\frac{n - 2}{3}$.

Summing the contributions of all node types:
\begin{equation*}
    \frac{n + 1}{3} + \frac{n + 1}{6} + \frac{n + 1}{6} + \frac{n - 2}{3} = n
\end{equation*}
as expected.

% III.7.4.
\topicblock{III.7.4}{Symbolic inclusion-exclusion.}

To make the relation $\mathcal{Q} \cong \mathcal{U} \star \mathcal{P}$ concrete, let us list the elements of $\mathcal{Q}_{3}$.
Here $\mathcal{U}$ records the distinguished fixed points, while $\mathcal{P}$ records the
permutation induced on the remaining labels. Thus an object of
$\mathcal{U}_k\star\mathcal{P}_{3-k}$ is obtained by choosing $k$ labels to be underlined fixed
points, and then permuting the remaining $3-k$ labels.
\begin{equation*}
    \begin{aligned}
        \mathcal{U} &= \{ \emptyset, \{1\}, \{1, 2\}, \{1, 2, 3\}, \cdots \} \\
        \mathcal{P} &= \{ \emptyset, (1), (1,2), (2,1), (1,2,3), (1,3,2), (2,1,3), (2,3,1), (3,1,2), (3,2,1), \cdots \}
    \end{aligned}
\end{equation*}
Pick an element of $\mathcal{U}$ of size $k$ and an element of $\mathcal{P}$ of size $3 - k$, with $0 \leq k \leq 3$.

\subhead{$\mathcal{U}_{0} \star \mathcal{P}_{3}$}
\begin{itemize}
    \item $\emptyset \star (1,2,3)$: $\binom{3}{0} = 1$ way of relabelling
        \begin{itemize}
            \item $(\emptyset, (1,2,3)) \leftrightarrow (1,2,3)$
        \end{itemize}
    \item $\emptyset \star (1,3,2)$: $\binom{3}{0} = 1$ way of relabelling
        \begin{itemize}
            \item $(\emptyset, (1,3,2)) \leftrightarrow (1,3,2)$
        \end{itemize}
    \item $\emptyset \star (2,1,3)$: $\binom{3}{0} = 1$ way of relabelling
        \begin{itemize}
            \item $(\emptyset, (2,1,3)) \leftrightarrow (2,1,3)$
        \end{itemize}
    \item $\emptyset \star (2,3,1)$: $\binom{3}{0} = 1$ way of relabelling
        \begin{itemize}
            \item $(\emptyset, (2,3,1)) \leftrightarrow (2,3,1)$
        \end{itemize}
    \item $\emptyset \star (3,1,2)$: $\binom{3}{0} = 1$ way of relabelling
        \begin{itemize}
            \item $(\emptyset, (3,1,2)) \leftrightarrow (3,1,2)$
        \end{itemize}
    \item $\emptyset \star (3,2,1)$: $\binom{3}{0} = 1$ way of relabelling
        \begin{itemize}
            \item $(\emptyset, (3,2,1)) \leftrightarrow (3,2,1)$
        \end{itemize}
\end{itemize}

\subhead{$\mathcal{U}_{1} \star \mathcal{P}_{2}$}
\begin{itemize}
    \item $\{1\} \star (1,2)$: $\binom{3}{1} = 3$ ways of relabelling
        \begin{itemize}
            \item $(\{1\}, (2,3)) \leftrightarrow (\underline{1},2,3)$
            \item $(\{2\}, (1,3)) \leftrightarrow (1,\underline{2},3)$
            \item $(\{3\}, (1,2)) \leftrightarrow (1,2,\underline{3})$
        \end{itemize}
    \item $\{1\} \star (2,1)$: $\binom{3}{1} = 3$ ways of relabelling
        \begin{itemize}
            \item $(\{1\}, (3,2)) \leftrightarrow (\underline{1},3,2)$
            \item $(\{2\}, (3,1)) \leftrightarrow (3,\underline{2},1)$
            \item $(\{3\}, (2,1)) \leftrightarrow (2,1,\underline{3})$
        \end{itemize}
\end{itemize}

\subhead{$\mathcal{U}_{2} \star \mathcal{P}_{1}$}
\begin{itemize}
    \item $\{1,2\} \star (1)$: $\binom{3}{2} = 3$ ways of relabelling
        \begin{itemize}
            \item $(\{1,2\}, (3)) \leftrightarrow (\underline{1},\underline{2},3)$
            \item $(\{1,3\}, (2)) \leftrightarrow (\underline{1},2,\underline{3})$
            \item $(\{2,3\}, (1)) \leftrightarrow (1,\underline{2},\underline{3})$
        \end{itemize}
\end{itemize}

\subhead{$\mathcal{U}_{3} \star \mathcal{P}_{0}$}
\begin{itemize}
    \item $\{1,2,3\} \star \emptyset$: $\binom{3}{3} = 1$ way of relabelling
        \begin{itemize}
            \item $(\{1,2,3\}, \emptyset) \leftrightarrow (\underline{1},\underline{2},\underline{3})$
        \end{itemize}
\end{itemize}

The count also matches:
\begin{equation*}
    |\mathcal{Q}_3|
        = \sum_{k=0}^{3}\binom{3}{k}(3-k)!
        = 6+6+3+1
        = 16.
\end{equation*}

% Example III.25.
\exampleblock{III.25}{Rises and ascending runs in permutations.}

\subhead{Rises}

For rises, inclusion--exclusion is applied to permutations with distinguished rises. Consecutive
distinguished rises form a cluster. A cluster with $m$ letters contains $m-1$ distinguished
rises, so its EGF is
\begin{equation*}
    C(z,v)=\sum_{m\geq 2}v^{m-1}\frac{z^m}{m!}
        = \frac{e^{zv}-1-zv}{v}.
\end{equation*}
A permutation with distinguished rises is then a sequence of isolated atoms and clusters:
\begin{equation*}
    B(z,v)=\frac{1}{1-z-C(z,v)}
        = \frac{v}{v+1-e^{zv}}.
\end{equation*}
If $A(z,u)$ marks all rises, then $B(z,v)=A(z,1+v)$, hence
\begin{equation*}
    A(z,u)=B(z,u-1)=\frac{u-1}{u-e^{z(u-1)}}.
\end{equation*}

\stephead{1}{Expand around $z = 0$.}
First expand $A(z, u)$ as a power series in $z$ around $0$.
\begin{equation*}
    A(z, u) = \frac{u - 1}{u - e^{z(u - 1)}}
\end{equation*}
The denominator is:
\begin{equation*}
    \begin{aligned}
        u - e^{z(u - 1)}
        &= u - \left[1 + (u - 1)z + \frac{(u - 1)^{2}}{2}z^{2} + \frac{(u - 1)^{3}}{6}z^{3} + \frac{(u - 1)^{4}}{24}z^{4} + O(z^{5})\right] \\
        &= (u - 1) - (u - 1)z - \frac{(u - 1)^{2}}{2}z^{2} - \frac{(u - 1)^{3}}{6}z^{3} - \frac{(u - 1)^{4}}{24}z^{4} + O(z^{5}) \\
        &= (u - 1)\left[1 - z - \frac{u - 1}{2} z^{2} - \frac{(u - 1)^{2}}{6}z^{3} - \frac{(u - 1)^{3}}{24}z^{4} + O(z^{5})\right]
    \end{aligned}
\end{equation*}
Hence,
\begin{equation*}
    \begin{aligned}
        A(z, u)
        &= \left(1 - z - \frac{u - 1}{2} z^{2} - \frac{(u - 1)^{2}}{6}z^{3} - \frac{(u - 1)^{3}}{24}z^{4} + O(z^{5})\right)^{-1} \\
        &= \left(1 - z - \frac{u - 1}{2} z^{2} - \frac{(u - 1)^{2}}{6}z^{3} - \frac{(u - 1)^{3}}{24}z^{4}\right)^{-1} + O(z^{5})
    \end{aligned}
\end{equation*}
Set
\begin{equation*}
    w = z + \frac{u - 1}{2} z^{2} + \frac{(u - 1)^{2}}{6}z^{3} + \frac{(u - 1)^{3}}{24}z^{4}
\end{equation*}
Compute powers of $w$ up to order $z^{4}$:
\begin{equation*}
    \begin{aligned}
        w^{2}
        &= \left(z + \frac{u - 1}{2} z^{2} + \frac{(u - 1)^{2}}{6}z^{3}\right)^{2} \\
        &= z^{2} + \left(\frac{u - 1}{2} z^{2}\right)^{2} + 2z \frac{u - 1}{2} z^{2} + 2z \frac{(u - 1)^{2}}{6}z^{3} \\
        &= z^{2} + (u - 1)z^{3} + \frac{7}{12}(u - 1)^{2}z^{4}
    \end{aligned}
\end{equation*}
\begin{equation*}
    w^{3} = \left(z + \frac{u - 1}{2} z^{2}\right)^{3} = z^{3} + 3z^{2} \frac{u - 1}{2} z^{2} = z^{3} + \frac{3}{2}(u - 1)z^{4}
\end{equation*}
\begin{equation*}
    w^{4} = z^{4}
\end{equation*}
Then
\begin{equation*}
    \begin{aligned}
        A(z, u)
        &= 1 + z + \left(\frac{u - 1}{2} + 1\right)z^{2} + \left(\frac{(u - 1)^{2}}{6} + (u - 1) + 1\right)z^{3} \\
        &\quad + \left(\frac{(u - 1)^{3}}{24} + \frac{7}{12}(u - 1)^{2} + \frac{3}{2}(u - 1) + 1\right)z^{4} + O(z^{5}) \\
        &= 1 + z + (u + 1)\frac{z^{2}}{2} + (u^{2} + 4u + 1)\frac{z^{3}}{6} + (u^{3} + 11u^{2} + 11u + 1)\frac{z^{4}}{24} + \cdots
    \end{aligned}
\end{equation*}

\stephead{2}{Expand around $u = 1$.}
Now expand $A(z, u)$ in powers of $u$ around $u = 1$. Set $v = u - 1$.
\begin{equation*}
    A(z, u) = \frac{v}{(v + 1) - e^{zv}}
\end{equation*}
The denominator
\begin{equation*}
    v + 1 - e^{zv} = v + 1 - \left(1 + zv + \frac{z^{2}}{2}v^{2} + \frac{z^{3}}{6}v^{3} + O(v^{4})\right)
        = (1 - z)v - \frac{z^{2}}{2}v^{2} - \frac{z^{3}}{6}v^{3} + O(v^{4})
\end{equation*}
Hence,
\begin{equation*}
    \begin{aligned}
        A(z, u)
        &= \left(1 - z - \frac{z^{2}}{2}v - \frac{z^{3}}{6}v^{2} + O(v^{3})\right)^{-1} \\
        &= \frac{1}{1 - z} \left(1 - \frac{z^{2}}{2(1 - z)}v - \frac{z^{3}}{6(1 - z)}v^{2} + O(v^{3})\right)^{-1} \\
        &= \frac{1}{1 - z} \left(1 - \frac{z^{2}}{2(1 - z)}v - \frac{z^{3}}{6(1 - z)}v^{2}\right)^{-1} + O(v^{3}) \\
        &= \frac{1}{1 - z} \left(1 + \frac{z^{2}}{2(1 - z)}v + \frac{z^{3}}{6(1 - z)}v^{2} + \left(\frac{z^{2}}{2(1 - z)}v\right)^{2}\right) + O(v^{3}) \\
        &= \frac{1}{1 - z} + \frac{z^{2}}{2(1 - z)^{2}}v + \frac{2z^{3} + z^{4}}{6(1 - z)^{3}} \frac{v^{2}}{2!} + \cdots
    \end{aligned}
\end{equation*}
\stephead{3}{Extract the mean and variance.}
In the expansion around $u=1$, the coefficient of $(u-1)$ is the cumulative GF for the
first factorial moment, while the coefficient of $(u-1)^2/2!$ is the cumulative GF for the
second factorial moment.
Extracting coefficients gives:
\begin{equation*}
    [z^{n}] \frac{z^{2}}{2(1 - z)^{2}} = \frac{1}{2} [z^{n - 2}] \sum_{m = 0}^{\infty}(m + 1)z^{m} = \frac{n - 1}{2}
\end{equation*}
\begin{equation*}
    [z^{n}] \frac{2z^{3} + z^{4}}{6(1 - z)^{3}}
        = \frac{1}{3} [z^{n - 3}] \sum_{m = 0}^{\infty} \binom{m + 2}{2} z^{m} + \frac{1}{6} [z^{n - 4}] \sum_{m = 0}^{\infty} \binom{m + 2}{2} z^{m}
        = \frac{3n^{2} - 11n + 10}{12}
\end{equation*}
The mean of the number of rises in a random permutation of size $n$ is $\frac{n - 1}{2}$, and for $n\geq 2$ the variance is:
\begin{equation*}
    \frac{3n^{2} - 11n + 10}{12} + \frac{n - 1}{2} - \left(\frac{n - 1}{2} \right)^{2} = \frac{n + 1}{12}
\end{equation*}

\subhead{Ascending Runs}

\stephead{1}{Fix the indexing first.}
Here it is safer not to use the word ``length'' too casually. I will say that an ascending run
has size $r$ if it consists of $r$ consecutive entries
\[
    \sigma_i < \sigma_{i+1} < \cdots < \sigma_{i+r-1}.
\]
Thus a rise has size $2$, but it has one adjacent comparison. With this convention, a cluster
of overlapping distinguished ascending runs has OGF
\begin{equation*}
    I(z,v)=\frac{v z^r}{1-v(z+z^2+\cdots+z^{r-1})}.
\end{equation*}
The numerator $z^r$ gives the first distinguished run. Each new overlapping run may start
$1,2,\ldots,r-1$ positions to the right of the previous one, which explains the denominator.
After the ordinary-to-exponential transform, permutations with distinguished runs are counted by
\begin{equation*}
    B(z,v)=\frac{1}{1-z-\widehat{I}(z,v)},\qquad
    A(z,u)=B(z,u-1).
\end{equation*}

\stephead{2}{Check the partial fraction statement.}
The rational function $I(z,v)$ is improper: the numerator has degree $r$, while the denominator
has degree $r-1$. For $r\geq 3$, polynomial division gives
\begin{equation*}
    I(z,v)
        =1-z+\frac{(v+1)z-1}{1-v(z+z^2+\cdots+z^{r-1})}.
\end{equation*}
Therefore the transformed form should be
\begin{equation*}
    \widehat{I}(z,v)
        =1-z+\sum_{j=1}^{r-1}c_j(v)e^{\omega_j(v)z},
\end{equation*}
where the $\omega_j(v)$ are the roots of
\[
    \omega^{r-1}=v(1+\omega+\cdots+\omega^{r-2}).
\]
The factor $(1-z)(v+1)$ printed in the book does not match this division. The special case
$r=2$ is also a good sanity check: then
\[
    I(z,v)=\frac{v z^2}{1-vz}
        =-z-\frac{1}{v}+\frac{1/v}{1-vz},
\]
so
\[
    \widehat{I}(z,v)=\frac{e^{vz}-1-vz}{v},
\]
which is exactly the cluster series used above for rises.

\stephead{3}{Estimate the moments.}
Let $X_{n,r}$ be the number of ascending runs of size $r$ in a random permutation of size $n$.
There are $n-r+1$ possible windows, and a fixed window of $r$ entries is increasing with
probability $1/r!$. Hence, exactly,
\begin{equation*}
    \mathbb{E}X_{n,r}=\frac{n-r+1}{r!}.
\end{equation*}
For the variance, write $X_{n,r}=\sum_i Y_i$, where $Y_i$ marks the window starting at $i$.
Disjoint windows are independent as far as their internal relative orders are concerned. If two
windows start $d$ positions apart, with $1\leq d\leq r-1$, then both are increasing iff the whole
block of $r+d$ entries is increasing. Thus
\begin{equation*}
    \mathbb{P}(Y_i=Y_{i+d}=1)=\frac{1}{(r+d)!}.
\end{equation*}
Consequently,
\begin{equation*}
    \begin{aligned}
    \mathrm{Var}(X_{n,r})
        &=(n-r+1)\frac{1}{r!}\left(1-\frac{1}{r!}\right) \\
        &\quad
        +2\sum_{d=1}^{r-1}(n-r+1-d)
            \left(\frac{1}{(r+d)!}-\frac{1}{(r!)^2}\right).
    \end{aligned}
\end{equation*}
In particular $\mathrm{Var}(X_{n,r})=O(n/r!)$ in the threshold range we care about. Therefore
\begin{equation*}
    \frac{\mathrm{Var}(X_{n,r})}{(\mathbb{E}X_{n,r})^2}
        =O\left(\frac{r!}{n}\right),
\end{equation*}
so Chebyshev gives concentration as soon as $r!=o(n)$. This is the natural condition for runs
of size $r$. In the book's later wording, the mean is written as
\begin{equation*}
    \frac{n-\ell+1}{\ell!},
\end{equation*}
which corresponds to our $r=\ell$. The displayed condition $(\ell+1)!=o(n)$ is stronger than
what this variance estimate needs; it is not the sharp threshold condition suggested by the
preceding mean estimate.

% Note III.36.
\noteblock{III.36}{Permutations without $\ell$-ascending runs.}

Here the book uses $\ell$ in the comparison-length sense: an $\ell$-ascending run has
$\ell+1$ entries. To avoid mixing this with the size convention above, set
$m=\ell+1$. Then excluding $\ell$-ascending runs is the same as excluding increasing
windows of size $m$.

Consider the denominator of $\left.A(z, u)\right|_{u = 0}$ before applying the
ordinary-to-exponential transform. From the cluster OGF with $r=m$ and $v=-1$,
\begin{equation*}
    1-z-I(z,-1)
        =1-z+\frac{z^m}{1+z+\cdots+z^{m-1}}
        =\frac{1}{1+z+\cdots+z^{m-1}}
        =\frac{1-z}{1-z^m}.
\end{equation*}
To prepare for the transform $(1-\omega z)^{-1}\mapsto e^{\omega z}$, decompose
$\frac{1-z}{1-z^m}$ using the $m$-th roots of unity. Let
$\omega_k=e^{2\pi i k/m}$ for $k=0,1,\ldots,m-1$.
The standard partial fraction expansion is:
\begin{equation*}
    \frac{1}{1-z^m}
        =\frac{1}{m}\sum_{k=0}^{m-1}\frac{1}{1-\omega_k z}.
\end{equation*}
Now apply the ordinary-to-exponential transform:
\begin{equation*}
    \begin{aligned}
        \text{EGF Denominator}
        &=\widehat{\frac{1-z}{1-z^m}} \\
        &=\widehat{\frac{1}{1-z^m}}
            -\widehat{\frac{z}{1-z^m}} \\
        &=\widehat{\frac{1}{1-z^m}}
            -\int_0^z \widehat{\frac{1}{1-t^m}}\,\dd t \\
        &=\frac{1}{m}\sum_{k=0}^{m-1}e^{\omega_k z}
            -\int_0^z\frac{1}{m}\sum_{k=0}^{m-1}e^{\omega_k t}\,\dd t \\
        &=\sum_{i\geq 0}\frac{z^{mi}}{(mi)!}
            -\sum_{i\geq 0}\frac{z^{mi+1}}{(mi+1)!}.
    \end{aligned}
\end{equation*}
At this point the root-of-unity filter has done its job:
$\sum_{k=0}^{m-1}\omega_k^n$ vanishes unless $m\mid n$, in which case it equals $m$.
The other small trick is that multiplying by $z$ in the ordinary series becomes integration
after passing to exponential generating functions.

Therefore the EGF of permutations without $\ell$-ascending runs is
\begin{equation*}
    \left(
        \sum_{i\geq 0}\frac{z^{(\ell+1)i}}{((\ell+1)i)!}
        -\sum_{i\geq 0}\frac{z^{(\ell+1)i+1}}{((\ell+1)i+1)!}
    \right)^{-1}.
\end{equation*}

For orientation, here are the first few EGFs when $m = 2,3,4$:
\begin{equation*}
    \begin{aligned}
        \text{A.R. of length 1} \quad \left(\sum_{i \geq 0} \frac{z^{2i}}{(2i)!} - \frac{z^{2i + 1}}{(2i + 1)!}\right)^{-1}
            &= 1 + z + \frac{z^{2}}{2!} + \frac{z^{3}}{3!} + \frac{z^{4}}{4!} + O(z^{5}) \\
        \text{A.R. of length 2} \quad \left(\sum_{i \geq 0} \frac{z^{3i}}{(3i)!} - \frac{z^{3i + 1}}{(3i + 1)!}\right)^{-1}
            &= 1 + z + 2 \frac{z^{2}}{2!} + 5 \frac{z^{3}}{3!} + 17 \frac{z^{4}}{4!} + O(z^{5}) \\
        \text{A.R. of length 3} \quad \left(\sum_{i \geq 0} \frac{z^{4i}}{(4i)!} - \frac{z^{4i + 1}}{(4i + 1)!}\right)^{-1}
            &= 1 + z + \frac{z^{2}}{2!} + \frac{z^{3}}{3!} + 23 \frac{z^{4}}{4!} + O(z^{5})
    \end{aligned}
\end{equation*}
For example, there are 17 permutations in $\mathcal{P}_4$ without ascending runs of length 2:
1324, 1423, 1432, 2143, 2314,
2413, 2431, 3142, 3214, 3241, 3412, 3421, 4132, 4213, 4231, 4312, 4321.

% Note III.41.
\noteblock{III.41}{Patterns in trees II.}

Let $\mathfrak{t}$ be the fixed non-empty pattern, with $m=|\mathfrak{t}|$. For a non-empty
binary tree $\tau=(\tau_0,\tau_1)$, each occurrence of $\mathfrak{t}$ is accounted for in
one of three places: the left subtree, the right subtree, or the root itself when
$\tau=\mathfrak{t}$. Thus
\begin{equation*}
    \omega[\tau]=\omega[\tau_0]+\omega[\tau_1]+\llbracket \tau=\mathfrak{t}\rrbracket,
    \qquad \omega[\emptyset]=0.
\end{equation*}
The only small trick is the following identity:
\begin{equation*}
    u^{\llbracket \tau = \mathfrak{t} \rrbracket}u^{\omega[\tau_{0}]}u^{\omega[\tau_{1}]}
        = u^{\omega[\tau_{0}]}u^{\omega[\tau_{1}]} + \llbracket \tau = \mathfrak{t} \rrbracket \cdot (u - 1)
\end{equation*}
Indeed, there are two cases.
\begin{itemize}
    \item $\tau \neq \mathfrak{t}$:
        $\text{LHS} = u^{0}u^{\omega[\tau_{0}]}u^{\omega[\tau_{1}]} = u^{\omega[\tau_{0}]}u^{\omega[\tau_{1}]}$;
        $\text{RHS} = u^{\omega[\tau_{0}]}u^{\omega[\tau_{1}]} + 0 \cdot (u - 1) = u^{\omega[\tau_{0}]}u^{\omega[\tau_{1}]}$
    \item $\tau = \mathfrak{t}$:
        the two proper subtrees of $\mathfrak{t}$ are smaller than $\mathfrak{t}$, so
        $\omega[\tau_0]=\omega[\tau_1]=0$. Hence
        $\text{LHS}=u$ and $\text{RHS}=1+(u-1)=u$.
\end{itemize}
Now sum over all binary trees $\tau$ to get the bivariate generating function:
\begin{equation*}
    \begin{aligned}
        B(z, u)
        &= \sum_{\tau} z^{|\tau|}u^{\omega[\tau]} \\
        &= z^{|\emptyset|}u^{\omega[\emptyset]} + \sum_{\tau \neq \emptyset} z^{|\tau|}u^{\omega[\tau]} \\
        &= z^{0}u^{0} + \sum_{\tau \neq \emptyset} z^{|\tau|}u^{\omega[\tau]} \\
        &= 1 + \sum_{\tau \neq \emptyset} z^{|\tau|}
            \left(u^{\omega[\tau_{0}]}u^{\omega[\tau_{1}]} + \llbracket \tau = \mathfrak{t} \rrbracket \cdot (u - 1)\right) \\
        &= 1 + \sum_{\tau_{0}, \tau_{1}}z^{1 + |\tau_{0}| + |\tau_{1}|}u^{\omega[\tau_{0}]}u^{\omega[\tau_{1}]}
            + \sum_{\tau \neq \emptyset} z^{|\tau|}\llbracket \tau = \mathfrak{t} \rrbracket \cdot (u - 1) \\
        &= 1 + z \sum_{\tau_{0}, \tau_{1}}z^{|\tau_{0}|}u^{\omega[\tau_{0}]} \cdot z^{|\tau_{1}|}u^{\omega[\tau_{1}]}
            + z^{|\mathfrak{t}|} \cdot (u - 1) \\
        &= 1 + zB(z, u)^{2} + z^{m}(u - 1)
    \end{aligned}
\end{equation*}
Solving the quadratic and choosing the branch with $B(0,u)=1$ gives
\begin{equation*}
    B(z,u)
        =\frac{1-\sqrt{1-4z-4(u-1)z^{m+1}}}{2z}.
\end{equation*}

\section{Extremal parameters}
% Note III.8.3.
\topicblock{III.8.3}{Averages and moments.}
Let $f^{[h]}(z)$ be the GF of objects whose extremal parameter is at most $h$, and set
$f^{[-1]}(z)=0$. Then $f^{[h]}(z)-f^{[h-1]}(z)$ counts objects whose parameter is exactly $h$.
To see why the first sum rearranges into the second, use a discrete summation-by-parts trick.
The rearrangement is coefficientwise: for a fixed object size, only finitely many values of $h$
can contribute, so no analytic convergence issue is being hidden here.
Write the linear factor $h$ as
\begin{equation*}
    h = \sum_{j = 1}^{h}1
\end{equation*}
Substituting this produces a double sum:
\begin{equation*}
    \Xi(z) = \sum_{h = 0}^{\infty} \left(\sum_{j = 1}^{h}1\right)\left[f^{[h]}(z) - f^{[h-1]}(z)\right]
        = \sum_{h = 1}^{\infty} \sum_{j = 1}^{h} \left[f^{[h]}(z) - f^{[h - 1]}(z)\right]
\end{equation*}
The summation region is $1 \leq j \leq h < \infty$. Reverse the order: then $j$ runs from $1$ to $\infty$, while $h$ runs from $j$ to $\infty$.
\begin{equation*}
    \Xi(z) = \sum_{j = 1}^{\infty} \sum_{h = j}^{\infty} \left[f^{[h]}(z) - f^{[h - 1]}(z)\right]
\end{equation*}
The inner sum over $h$ now telescopes:
\begin{equation*}
    \sum_{h = j}^{\infty} \left[f^{[h]}(z) - f^{[h - 1]}(z)\right] = \lim_{N \to \infty} \sum_{h = j}^{N} \left[ f^{[h]}(z) - f^{[h - 1]}(z)\right]
\end{equation*}
Expanding the finite sum makes the cancellation explicit:
\begin{equation*}
    \lim_{N \to \infty} \left(\left[f^{[j]} - f^{[j - 1]}\right] + \left[f^{[j + 1]} - f^{[j]}\right] + \dots + \left[f^{[N]} - f^{[N - 1]}\right]\right)
        = \lim_{N \to \infty} \left(f^{[N]}(z) - f^{[j - 1]}(z)\right)
\end{equation*}
Because $f^{[N]}(z)$ counts objects with parameter at most $N$, letting $N \to \infty$ recovers the unconstrained generating function $f(z)$.
So the inner sum collapses to
\begin{equation*}
    f(z) - f^{[j - 1]}(z)
\end{equation*}
Putting this back into the outer sum,
\begin{equation*}
    \Xi(z) = \sum_{j = 1}^{\infty} \left[f(z) - f^{[j - 1]}(z)\right]
\end{equation*}
Finally shift the index by $h = j - 1$:
\begin{equation*}
    \Xi(z) = \sum_{h = 0}^{\infty} \left[f(z) - f^{[h]}(z)\right]
\end{equation*}

\section{Perspective}
