% !TEX root = main.tex
\chapter{COMBINATORIAL STRUCTURES AND ORDINARY GENERATING FUNCTIONS}
\label{chap:combinatorial structures and ordinary generating functions}

\section{Symbolic enumeration methods}
% Note I.2
\noteblock{I.2}{Unimodal permutations.}

A unimodal permutation rises to a single summit, namely $n$, and then comes back down.
So the whole shape is: an increasing run, the peak $n$, and then a decreasing run.
This class admits the symbolic specification
\begin{equation*}
    \mathcal{P}^{\mathcal{U}} = \mathrm{S}\textsc{et}(\mathcal{Z}) \star \mathcal{Z}^{\blacksquare} \star \mathrm{S}\textsc{et}(\mathcal{Z})
\end{equation*}
Accordingly, its exponential generating function is
\begin{equation*}
    P^{U}(z) = \int_{0}^{z}\left(e^{t} \cdot \frac{\dd}{\dd t} t \cdot e^{t}\right)\dd t = \frac{1}{2}(e^{2z} - 1)
\end{equation*}
and therefore
\begin{equation*}
    P^{U}_{n} = n![z^{n}] \frac{1}{2} e^{2z} = n!\frac{1}{2} \frac{2^{n}}{n!} = 2^{n - 1}.
\end{equation*}

There is also a one-line counting proof. If $n$ sits in position $k$, then we only need to choose which
$k-1$ of the remaining $n-1$ entries go to its left; once that choice is made, the left-hand side must be
in increasing order and the right-hand side in decreasing order. Hence
\begin{equation*}
    P^{U}_{n} = \sum_{k = 1}^{n} \binom{n - 1}{k - 1}
    = \sum_{j = 0}^{n - 1} \binom{n - 1}{j}
    = 2^{n - 1}.
\end{equation*}

\section{Admissible constructions and specifications}

\section{Integer compositions and partitions}
% Note I.16
\noteblock{I.16}{Partitions with summands bounded in number and size.}

Consider partitions of size $n$ with at most $k$ summands, each bounded by $l$. The corresponding counting formula is
\begin{equation} \label{eq:I.16.1}
    [z^{n}] \frac{(1 - z)(1 - z^{2}) \cdots (1 - z^{k + l})}{((1 - z)(1 - z^{2}) \cdots (1 - z^{k})) \cdot ((1 - z)(1 - z^{2}) \cdots (1 - z^{l}))}
\end{equation}
Let $f(k, l)$ denote the rational function inside \eqref{eq:I.16.1}, so that the coefficient of $z^{n}$ counts the admissible partitions of $n$.
The natural split is by the largest part: it is either strictly smaller than $l$, or exactly $l$. Thus
\begin{equation} \label{eq:I.16.2}
    f(k, l) = f(k, l - 1) + z^{l}f(k - 1, l)
\end{equation}
To package the dependence on $l$, write
\begin{equation*}
    F_{k}(x) = \sum_{l = 0}^{\infty} f(k, l)x^{l}
\end{equation*}
For the base case $k = 0$, only the empty partition survives, so $f(0,l) = 1$ for all $l \geq 0$. Hence
\begin{equation*}
    F_{0}(x) = \sum_{l = 0}^{\infty}1 \cdot x^{l} = \frac{1}{1 - x}
\end{equation*}
Now multiply \eqref{eq:I.16.2} by $x^{l}$ and sum over $l \geq 1$:
\begin{equation*}
    \sum_{l = 1}^{\infty} f(k, l)x^{l} - \sum_{l = 1}^{\infty} f(k, l - 1)x^{l} = \sum_{l = 1}^{\infty} f(k - 1, l)(zx)^{l}
\end{equation*}
\begin{itemize}
    \item The first term simplifies to $F_{k}(x) - 1$, since $f(k, 0) = 1$.
    \item The second term is simply $x F_{k}(x)$.
    \item The right-hand side yields $F_{k - 1}(zx) - 1$.
\end{itemize}
Substituting these pieces back in gives
\begin{equation*}
    (1 - x)F_{k}(x) - 1 = F_{k - 1}(zx) - 1 \implies F_{k}(x) = \frac{1}{1 - x} F_{k - 1}(zx)
\end{equation*}
Iterating the relation gives
\begin{equation*}
    \begin{aligned}
        F_{k}(x)
        &= \frac{1}{1 - x} F_{k - 1}(zx) \\
        &= \frac{1}{1 - x} \frac{1}{1 - zx} F_{k - 2}(z^{2}x) \\
        &= \frac{1}{(1 - x)(1 - zx) \cdots (1 - z^{k - 1}x)}F_{0}(z^{k}x)
    \end{aligned}
\end{equation*}
Since $F_{0}(x) = \frac{1}{1 - x}$, we have $F_{0}(z^{k}x) = \frac{1}{1 - z^{k}x}$. Thus
\begin{equation*}
    F_{k}(x) = \sum_{l = 0}^{\infty} f(k, l)x^{l} = \frac{1}{(1 - x)(1 - zx) \cdots (1 - z^{k}x)}
\end{equation*}
To identify this generating function more explicitly, we now decompose it into partial fractions:
\begin{equation*}
    \frac{1}{\prod_{i = 0}^{k}(1 - z^{i}x)} = \sum_{j = 0}^{k} \frac{A_{j}}{1 - z^{j}x}
\end{equation*}
To isolate $A_{j}$, multiply through by $(1 - z^{j}x)$ and then set $x = z^{-j}$:
\begin{equation*}
    \begin{aligned}
        A_{j}
        &= \left[\frac{1 - z^{j}x}{\prod_{i = 0}^{k}(1 - z^{i}x)}\right]_{x = z^{-j}} \\
        &= \frac{1}{\prod_{i = 0}^{j - 1}(1 - z^{i}z^{-j})\prod_{i = j + 1}^{k}(1 - z^{i}z^{-j})} \\
        &= \frac{(-1)^{j}z^{j(j + 1)/2}}{\prod_{i = 1}^{j}(1 - z^{i})\prod_{i = 1}^{k - j}(1 - z^{i})}
    \end{aligned}
\end{equation*}
The coefficient of $x^{l}$ in $F_{k}(x)$ is then read off term by term. For $\frac{A_{j}}{1 - z^{j}x}$, it is simply $A_{j}(z^{j})^{l}$.
\begin{equation*}
    \begin{aligned}
        f(k, l)
        &= \sum_{j = 0}^{k} A_{j}(z^{j})^{l} \\
        &= \sum_{j = 0}^{k} \frac{(-1)^{j}z^{j(j + 1)/2} \cdot z^{jl}}{\prod_{i = 1}^{j}(1 - z^{i})\prod_{i = 1}^{k - j}(1 - z^{i})} \\
        &= \frac{1}{\prod_{i = 1}^{k}(1 - z^{i})} \sum_{j = 0}^{k} \binom{k}{j}_{z}(-1)^{j}z^{j(j + 1)/2 + jl}
    \end{aligned}
\end{equation*}
where $\binom{k}{j}_{z}$ is the Gaussian binomial coefficient:
\begin{equation*}
    \begin{aligned}
        \binom{k}{j}_{z}
        &= \frac{\prod_{i = 1}^{k}(1 - z^{i})}{\prod_{i = 1}^{j}(1 - z^{i})\prod_{i = 1}^{k - j}(1 - z^{i})} \\
        &= \frac{(1 - z^{k})(1 - z^{k - 1}) \cdots (1 - z^{k - j + 1})}{(1 - z)(1 - z^{2}) \cdots (1 - z^{j})}
    \end{aligned}
\end{equation*}
A small rearrangement of the exponent gives
\begin{equation*}
    \frac{j(j + 1)}{2} + jl = \frac{j(j - 1)}{2} + j(l + 1)
\end{equation*}
so the sum becomes
\begin{equation*}
    f(k, l) = \frac{1}{\prod_{i = 1}^{k}(1 - z^{i})} \sum_{j = 0}^{k} \binom{k}{j}_{z}(-1)^{j}z^{j(j - 1)/2}(z^{l + 1})^{j}
\end{equation*}
The $q$-binomial theorem states that for a variable $X$:
\begin{equation} \label{eq:I.16.3}
    \prod_{i = 0}^{k - 1}(1 - z^{i}X) = \sum_{j = 0}^{k} \binom{k}{j}_{z}(-1)^{j}z^{j(j - 1)/2}X^{j}
\end{equation}
Substituting $X = z^{l + 1}$ into the theorem recovers \eqref{eq:I.16.1}:
\begin{equation*}
    \begin{aligned}
        f(k, l)
        &= \frac{(1 - z^{l + 1})(1 - z^{l + 2}) \cdots (1 - z^{l + k})}{(1 - z^{1})(1 - z^{2}) \cdots (1 - z^{k})} \\
        &= \frac{(1 - z)(1 - z^{2}) \cdots (1 - z^{k + l})}{((1 - z)(1 - z^{2}) \cdots (1 - z^{k})) \cdot ((1 - z)(1 - z^{2}) \cdots (1 - z^{l}))}
    \end{aligned}
\end{equation*}
At this point, the original product formula has been recovered. The rest is optional background, but it is pleasant enough to record: let us also justify \eqref{eq:I.16.3}. Set
\begin{equation*}
    P_{k}(X) = \prod_{i = 0}^{k - 1}(1 - z^{i}X)
\end{equation*}
We claim that $P_{k}(X) = \sum_{j = 0}^{k} \binom{k}{j}_{z}(-1)^{j}z^{j(j - 1)/2}X^{j}$.
The useful relation comes from comparing $P_k(zX)$ with $P_k(X)$:
\begin{equation*}
    P_{k}(zX) = \prod_{i = 0}^{k - 1}(1 - z^{i + 1}X) = \frac{1 - z^{k}X}{1 - X}P_{k}(X)
\end{equation*}
After rearranging, this becomes the functional equation
\begin{equation*}
    (1 - X)P_{k}(zX) = (1 - z^{k}X)P_{k}(X)
\end{equation*}
Let $P_{k}(X) = \sum_{j = 0}^{k} c_{j}X^{j}$. We substitute this power series into the functional equation:
\begin{equation*}
    (1 - X)\sum_{j = 0}^{k} c_{j}(zX)^{j} = (1 - z^{k}X)\sum_{j = 0}^{k} c_{j}X^{j}
\end{equation*}
Comparing coefficients of $X^{j}$ on both sides yields:
\begin{equation*}
    c_{j}z^{j} - c_{j - 1}z^{j - 1} = c_{j} - c_{j - 1}z^{k}
\end{equation*}
so
\begin{equation*}
    c_{j} = c_{j - 1} \frac{z^{j - 1} - z^{k}}{z^{j} - 1} = -c_{j - 1}z^{j - 1} \frac{1 - z^{k - j + 1}}{1 - z^{j}}
\end{equation*}
Starting from $c_{0} = P_{k}(0) = 1$, we iterate:
\begin{equation*}
    c_{1} = -c_{0}z^{0} \frac{1 - z^{k}}{1 - z^{1}}, \quad
    c_{2} = -c_{1}z^{1} \frac{1 - z^{k - 1}}{1 - z^{2}} = (-1)^{2}z^{0 + 1} \frac{(1 - z^{k})(1 - z^{k - 1})}{(1 - z^{1})(1 - z^{2})}
\end{equation*}
The exponent of $z$ is $0 + 1 + 2 + \cdots + (j - 1) = \frac{j(j - 1)}{2}$, and the remaining fraction is exactly the Gaussian binomial coefficient. Therefore,
\begin{equation*}
    c_{j} = (-1)^{j}z^{j(j - 1)/2} \binom{k}{j}_{z}
\end{equation*}
Plugging the coefficients back into the expansion, we obtain
\begin{equation*}
    \prod_{i = 0}^{k - 1}(1 - z^{i}X) = \sum_{j = 0}^{k} \binom{k}{j}_{z}(-1)^{j}z^{j(j - 1)/2}X^{j}
\end{equation*}
as claimed.

% Note I.17
\noteblock{I.17}{Systems of linear Diophantine inequalities.}

This note contains five counting problems. They share a common theme, but each has its own constraint set and can be solved independently.

\subhead{Problem 1}
\begin{equation*}
    \left\{x_{1}\geq 0,\; x_{2}\geq 2x_{1},\; x_{3}\geq 2x_{2},\; x_{4}\geq 2x_{3}\right\}
\end{equation*}
At first glance this system looks a little forbidding, but slack variables settle it quickly. Introduce non-negative variables
$y_{j} \in \mathbb{Z}_{\geq 0}$ to measure the excess above each lower bound.
\begin{align*}
    x_{1} &= y_{1} \\
    x_{2} &= 2x_{1} + y_{2} = 2y_{1} + y_{2} \\
    x_{3} &= 2x_{2} + y_{3} = 4y_{1} + 2y_{2} + y_{3} \\
    x_{4} &= 2x_{3} + y_{4} = 8y_{1} + 4y_{2} + 2y_{3} + y_{4}
\end{align*}
The total weight is
\begin{equation*}
    S = x_{1} + x_{2} + x_{3} + x_{4} = 15y_{1} + 7y_{2} + 3y_{3} + y_{4}.
\end{equation*}
The generating function is now immediate: the variables $y_{j}$ are independent and each ranges freely over
the non-negative integers, so the OGF $F(z)$ is the product of four one-variable OGFs:
\begin{itemize}
    \item For $15y_{1}: \sum_{y_{1} = 0}^{\infty}(z^{15})^{y_{1}} = \frac{1}{1 - z^{15}}$
    \item For $ 7y_{2}: \sum_{y_{2} = 0}^{\infty}(z^{ 7})^{y_{2}} = \frac{1}{1 - z^{ 7}}$
    \item For $ 3y_{3}: \sum_{y_{3} = 0}^{\infty}(z^{ 3})^{y_{3}} = \frac{1}{1 - z^{ 3}}$
    \item For $  y_{4}: \sum_{y_{4} = 0}^{\infty}(z^{ 1})^{y_{4}} = \frac{1}{1 - z^{ 1}}$
\end{itemize}
Multiplying them together gives the answer:
\begin{equation*}
    F(z) = \frac{1}{(1 -z)(1 -z^{3})(1 -z^{7})(1 -z^{15})}
\end{equation*}

\partsep
\subhead{Problem 2}
\begin{equation*}
    \left\{x_{1} + x_{2}\leq x_{3}\right\}
\end{equation*}
Here slack variables again turn the inequality into an equality with independent parts.
Rewrite the constraint $x_{3}\geq x_{1} + x_{2}$ by introducing a non-negative integer $y\geq 0$:
\begin{equation*}
    x_{3} = x_{1} + x_{2} + y
\end{equation*}
The total size is $S = x_{1} + x_{2} + x_{3} + x_{4} = 2x_{1} + 2x_{2} + y + x_{4}$.
Since the variables $x_{1}, x_{2}, y, x_{4}$ are independent and each ranges freely over the non-negative integers, the OGF is the product of
the corresponding one-variable generating functions:
\begin{itemize}
    \item For $2x_{1}: \frac{1}{1 - z^{2}}$
    \item For $2x_{2}: \frac{1}{1 - z^{2}}$
    \item For $y: \frac{1}{1 - z}$
    \item For $x_{4}: \frac{1}{1 - z}$
\end{itemize}
Multiplying these factors gives
\begin{equation*}
    F(z) = \frac{1}{(1 -z^{2})^{2}(1 -z)^{2}}
\end{equation*}

\partsep
\subhead{Problem 3}
\begin{equation*}
    \left\{x_{1} + x_{2}\geq x_{3}\right\}
\end{equation*}
Since the inequality does not suggest a single slack variable, the cleanest route is to work with the complement.
Let $U(z) = (1 - z)^{-4}$ be the OGF for four unrestricted variables. The universe of compositions splits into those with $x_{1} + x_{2} < x_{3}$ and those with $x_{1} + x_{2} \geq x_{3}$.
So we first count the complement class $x_{1} + x_{2} < x_{3}$ by setting $x_{3} = x_{1} + x_{2} + 1 + y$, where $y\geq 0$.
The total sum is
\begin{equation*}
    S = x_{1} + x_{2} + (x_{1} + x_{2} + 1 + y) + x_{4} = 2x_{1} + 2x_{2} + y + x_{4} + 1
\end{equation*}
The OGF for this complement, denoted by $C(z)$, is
\begin{equation*}
    C(z) = \frac{z}{(1 - z^{2})^{2}(1 - z)^{2}}
\end{equation*}
The OGF for our constraint is then
\begin{equation*}
    F(z) = U(z) - C(z) = \frac{1}{(1 - z)^{4}} - \frac{z}{(1 - z^{2})^{2}(1 - z)^{2}} = \frac{1 + z + z^{2}}{(1 - z)^{4}(1 + z)^{2}}
\end{equation*}

\partsep
\subhead{Problem 4}
\begin{equation*}
    \left\{x_{1} + x_{2}\leq x_{3} + x_{4}\right\}
\end{equation*}

To find the OGF for $\left\{x_{1} + x_{2}\leq x_{3} + x_{4}\right\}$, combine symmetry with a simple decomposition. Let $A = x_{1} + x_{2}$ and $B = x_{3} + x_{4}$.
We seek the OGF for compositions satisfying $A\leq B$. In the universe of all compositions
$(x_{1}, x_{2}, x_{3}, x_{4})$, there are three mutually exclusive cases:
$1.\; A < B,\; 2.\; A > B,\; 3.\; A = B$. By symmetry, cases 1 and 2 have the same OGF,
since $A$ and $B$ are built from the same number of variables with the same weights.
Denote this common OGF by $G_{<}(z)$, and let $G_{=}(z)$ be the OGF for case 3.
The total unrestricted OGF is $U(z) = \frac{1}{(1 - z)^{4}}$. Therefore,
\begin{equation*}
    U(z) = 2G_{<}(z) + G_{=}(z)
\end{equation*}
Our target OGF is $F(z) = G_{<}(z) + G_{=}(z)$. Solving the system gives:
\begin{equation*}
    F(z) = \frac{U(z) + G_{=}(z)}{2}
\end{equation*}
Now $G_{=}(z)$ is the OGF for compositions in which the sum of the first two variables equals
the sum of the last two.
\begin{itemize}
    \item The number of compositions with $x_{1} + x_{2} = n$ is:
        $[z^{n}] \frac{1}{(1 - z)^{2}} = n + 1$.
    \item The number of compositions with $x_{3} + x_{4} = n$ is the same.
    \item For the equality $x_{1} + x_{2} = x_{3} + x_{4}$ to hold, both sums must equal some $n\geq 0$.
        \begin{equation*}
            G_{=}(z) = \sum_{n = 0}^{\infty}(n + 1)^{2}z^{2n} = \frac{1 + z^{2}}{(1 - z^{2})^{3}}
        \end{equation*}
\end{itemize}
Substituting $U(z)$ and $G_{=}(z)$ into our formula yields:
\begin{equation*}
    F(z) = \frac{1}{2} \left[\frac{1}{(1 - z)^{4}} + \frac{1 + z^{2}}{(1 - z^{2})^{3}}\right] = \frac{1 + z + 2z^{2}}{(1 - z)^{4}(1 + z)^{3}}
\end{equation*}

\partsep
\subhead{Problem 5}
\begin{equation*}
    \left\{x_{1}\leq x_{2},\; x_{2}\geq x_{3},\; x_{3}\leq x_{4}\right\}
\end{equation*}
Because the inequalities alternate direction, the simple slack-variable chain for $x_{1}\leq x_{2}\leq x_{3}\leq x_{4}$ no longer applies. Instead, we use inclusion--exclusion.
We wish to enforce three conditions
\begin{itemize}
    \item $C_{1}: x_{1}\leq x_{2}$
    \item $C_{2}: x_{2}\geq x_{3}$
    \item $C_{3}: x_{3}\leq x_{4}$
\end{itemize}
It is easiest to keep the two outer inequalities first, and then subtract the cases where the middle constraint fails.
So first consider the set of compositions satisfying both $x_{1}\leq x_{2}$ and $x_{3}\leq x_{4}$.
The OGF for this set is easy to obtain because the two pairs are independent.
\begin{itemize}
    \item For $x_{1}\leq x_{2}$, set $x_{2} = x_{1} + y_{1}$. The total contribution is $2x_{1} + y_{1}$,
        so the OGF is $\frac{1}{(1 - z^{2})(1 - z)}$.
    \item For $x_{3}\leq x_{4}$, the OGF is identically $\frac{1}{(1 - z^{2})(1 - z)}$.
\end{itemize}
Since the two pairs are independent, we multiply their OGFs:
\begin{equation*}
    A(z) = \frac{1}{(1 - z^{2})^{2}(1 - z)^{2}}
\end{equation*}
Now we find the OGF for the violation $x_{1}\leq x_{2} < x_{3}\leq x_{4}$ by introducing slack variables
\begin{itemize}
    \item $x_{1} = y_{1}$
    \item $x_{2} = y_{1} + y_{2}$
    \item $x_{3} = y_{1} + y_{2} + 1 + y_{3}$
    \item $x_{4} = y_{1} + y_{2} + 1 + y_{3} + y_{4}$
\end{itemize}
Summing the four expressions gives
\begin{equation*}
    S = 4y_{1} + 3y_{2} + 2y_{3} + y_{4} + 2
\end{equation*}
The OGF for this violation is therefore
\begin{equation*}
    V(z) = \frac{z^{2}}{(1 - z^{4})(1 - z^{3})(1 - z^{2})(1 - z)}
\end{equation*}
The OGF for the zigzag constraint is the OGF for the two independent pairs minus the OGF for the cases where the peak at $x_{2}$ fails:
\begin{equation*}
    \begin{aligned}
        F(z)
        &= \frac{1}{(1 - z^{2})^{2}(1 - z)^{2}} - \frac{z^{2}}{(1 - z^{4})(1 - z^{3})(1 - z^{2})(1 - z)} \\
        &= \frac{1}{(1 - z)(1 - z^{2})}\left[\frac{1}{(1 - z)(1 - z^{2})} - \frac{z^{2}}{(1 - z^{3})(1 - z^{4})}\right]
    \end{aligned}
\end{equation*}

% Note I.18
\noteblock{I.18}{Odd versus distinct summands.}

This bijection looks clever at first and almost inevitable afterwards. We build a one-to-one correspondence between partitions into odd parts and partitions into distinct parts.
\begin{itemize}
    \item Odd $\rightarrow$ Distinct
        \begin{enumerate}
            \item Group equal odd parts together.
            \item If an odd part $k$ appears $m$ times, write $m$ in binary as
                    $m = 2^{a_{1}} + 2^{a_{2}} + \cdots$ with distinct powers of two.
            \item Replace those $m$ copies of $k$ by the distinct parts $2^{a_{1}} \cdot k,\;
                    2^{a_{2}} \cdot k,\; \dots$
        \end{enumerate}
    \item Distinct $\rightarrow$ Odd
        \begin{enumerate}
            \item Factor each part as $2^{a}\times \text{odd}$.
            \item Replace it by $2^{a}$ copies of that odd factor.
            \item Do this for every part; the result is a partition into odd parts.
            \item Sort the resulting parts to obtain a partition into odd parts.
        \end{enumerate}
\end{itemize}
Here is the map in action:
\begin{equation*}
    \begin{aligned}
        3 + 3 + 3 + 3 + 3 + 1 + 1
        &= 5\times 3 + 2\times 1 \\
        &= (2^{2} + 2^{0})\times 3 + 2^{1}\times 1 \\
        &= 4\times 3 + 1\times 3 + 2\times 1 \\
        &= 12 + 3 + 2
    \end{aligned}
\end{equation*}
\begin{equation*}
    \begin{aligned}
        12 + 3 + 2
        &= 4\times 3 + 3 + 2\times 1 \\
        &= (3 + 3 + 3 + 3) + 3 + (1 + 1) \\
        &= 3 + 3 + 3 + 3 + 3 + 1 + 1
    \end{aligned}
\end{equation*}

% Note I.19
\noteblock{I.19}{Euler’s pentagonal number theorem.}

To get a usable recurrence for the partition numbers, it is natural to begin with the reciprocal of $P(z)$:
\begin{equation*}
    A(z) = \prod_{n = 1}^{\infty}(1 - z^{n})
\end{equation*}
Euler's Pentagonal Number Theorem says that this product collapses to a remarkably sparse series:
\begin{equation*}
    \begin{aligned}
        A(z)
            &= \sum_{k = -\infty}^{\infty}(-1)^{k}z^{k(3k + 1)/2} \\
            &= \sum_{k = -\infty}^{\infty}(-1)^{k}z^{k(3k - 1)/2} \\
            &= 1 - z^{1} - z^{2} + z^{5} + z^{7} - z^{12} - z^{15} + \cdots
    \end{aligned}
\end{equation*}
The exponents $g_{k} = \frac{k(3k - 1)}{2}$ are the generalized pentagonal numbers. Since $A(z)$ is the reciprocal of $P(z)$, this gives
\begin{equation*}
    \sum_{n = 0}^{\infty} p(n)z^{n} \cdot \sum_{k = -\infty}^{\infty}(-1)^{k}z^{g_{k}} = 1
\end{equation*}
Comparing coefficients of $z^{n}$ now gives a recurrence that is genuinely useful in computation:
\begin{equation*}
    p(n) = p(n - 1) + p(n - 2) - p(n - 5) - p(n - 7) + p(n - 12) + p(n - 15) \cdots
\end{equation*}
A straightforward Python implementation follows:
\newpage
\begin{lstlisting}
def compute_partitions(N):
    P = [0] * (N + 1)
    P[0] = 1  # Base case: p(0) = 1

    for i in range(1, N + 1):
        k = 1
        while True:
            # Calculate pentagonal numbers for k and -k
            g_k_pos = k * (3 * k - 1) // 2
            g_k_neg = k * (3 * k + 1) // 2

            # Determine the sign: +1 for k=1,2, -1 for k=3,4, and so on.
            # This is exactly (-1)^(k-1).
            sign = 1 if k % 2 == 1 else -1

            # Add p(i - g_k_pos)
            if g_k_pos <= i:
                P[i] += sign * P[i - g_k_pos]
            else:
                break  # No more terms fit

            # Add p(i - g_k_neg)
            if g_k_neg <= i:
                P[i] += sign * P[i - g_k_neg]
            else:
                break

            k += 1
    return P
\end{lstlisting}

% Note I.20
\noteblock{I.20}{A digital surprise.}
\begin{equation*}
    \begin{aligned}
        \varphi
            &= (1 - 10^{-1})(1 - 10^{-2})(1 - 10^{-3})(1 - 10^{-4}) \cdots \\
            &= \left.\prod_{n = 1}^{\infty}(1 - z^{n})\right|_{z = .1} \\
            &= \left.1 - z^{1} - z^{2} + z^{5} + z^{7} - z^{12} - z^{15} + z^{22} + z^{26} - \cdots \right|_{z = .1} \\
            &= .\dot{9} - .1 - .01 + .00001 + .0000001 - .000000000001 - \cdots
    \end{aligned}
\end{equation*}
The partial sums are
\begin{itemize}
    \item $\varphi_{1} = .99999 99999 99999 99999 99999$
    \item $\varphi_{2} = .89999 99999 99999 99999 99999$
    \item $\varphi_{3} = .88999 99999 99999 99999 99999$
    \item $\varphi_{4} = .89000 99999 99999 99999 99999$
    \item $\varphi_{5} = .89001 00999 99999 99999 99999$
    \item $\varphi_{6} = .89001 00999 98999 99999 99999$
    \item $\varphi_{7} = .89001 00999 98998 99999 99999$
    \item $\varphi_{8} = .89001 00999 98999 00000 00999$
\end{itemize}
This also explains why only the digits $0,1,8,9$ appear. Start from $.999\ldots$.
Each new term is either $-10^{-m}$ or $+10^{-m}$ for some generalized pentagonal
exponent $m$, so at each step we modify only one decimal place together with the
possible carry or borrow to its left. A subtraction can only change a digit $9$ to $8$.
An addition can create a carry through a block of $9$'s, turning that block into $0$'s and
eventually producing a $1$; if the carry hits an $8$, that $8$ simply becomes $9$.
Thus the only possible digit changes are
\[
9 \to 8,\qquad 9 \to 0,\qquad 8 \to 9,\qquad 0 \to 1,
\]
and no other digit can ever be created.

% Note I.21
\noteblock{I.21}{Lattice points.}

We want the number of integer solutions to $\sum_{i = 1}^{d} x_{i}^{2} \le n^{2}$.
The natural tool here is the theta series. For one coordinate $x_{i} \in \mathbb{Z}$, the contribution to the squared radius is $x_{i}^{2}$, so the one-variable generating function is
\begin{equation*}
    \Theta(z) = \sum_{x \in \mathbb{Z}}z^{x^{2}} = 1 + 2 \sum_{n = 1}^{\infty} z^{n^{2}}
\end{equation*}
Because the coordinates are independent, the generating function for the total quantity $\sum x_{i}^{2}$ is simply the $d$th power:
\begin{equation*}
    S(z) = (\Theta(z))^{d} = \sum_{k = 0}^{\infty} r_{d}(k)z^{k}
\end{equation*}
Here $r_{d}(k) = [z^{k}]S(z)$ counts lattice points whose squared distance from the origin is
exactly $k$. The ball of radius $n$ asks for squared distance at most $n^{2}$, so we need the
cumulative sums of the coefficients $r_{d}(k)$ up to $n^{2}$. Generating functions encode
these cumulative sums by multiplying by $\frac{1}{1 - z}$:
\begin{equation*}
    \sum_{k = 0}^{\infty} \left(\sum_{j = 0}^{k} r_{d}(j)\right)z^{k} = \frac{1}{1 - z} S(z) = \frac{1}{1 - z}(\Theta(z))^{d}
\end{equation*}
The coefficient of $z^{n^{2}}$ is precisely the sum of all $r_{d}(j)$ with $j \leq n^{2}$, so
\begin{equation*}
    N(n, d) = [z^{n^{2}}] \frac{1}{1 - z}(\Theta(z))^{d}
\end{equation*}

\section{Words and regular languages}
% Note I.23
\noteblock{I.23}{Alice, Bob, and coding bounds.}

Alice wants to encode an arbitrary binary message into a binary word that avoids the forbidden pattern $11$.
If messages of length $n$ are encoded into admissible words of length at most $L$, then we need at least
\begin{equation*}
    2^{n} \leq \sum_{j = 0}^{L} C_{j},
\end{equation*}
where $C_{j}$ counts admissible codewords of length $j$.

By Proposition I.4, the class $\mathcal{C}$ of binary strings avoiding the pattern $11$ has generating function
\begin{equation*}
    C(z) = \frac{1 + z}{1 - z - z^{2}}
\end{equation*}
so $C_{j} = F_{j + 2}$, with $(F_{m})$ the Fibonacci numbers. Therefore
\begin{equation*}
    \sum_{j = 0}^{L} C_{j} = \sum_{j = 0}^{L} F_{j + 2} = F_{L + 4} - 2.
\end{equation*}
This already gives the exact counting bound
\begin{equation*}
    2^{n} \leq F_{L + 4} - 2.
\end{equation*}

If one prefers generating functions, the partial sums are encoded by
\begin{equation*}
    S(z) = \frac{C(z)}{1 - z} = \frac{1 + z}{(1 - z)(1 - z - z^{2})}.
\end{equation*}
Either way, the asymptotics are governed by the golden ratio $\varphi = \frac{1+\sqrt{5}}{2}$, since
\begin{equation*}
    F_{L + 4} \sim \frac{\varphi^{L + 4}}{\sqrt{5}}.
\end{equation*}
Thus, as $L \to \infty$,
\begin{equation*}
    \sum_{j = 0}^{L} C_{j} \sim A \cdot \varphi^{L}
\end{equation*}
for some constant $A > 0$. Taking base-2 logarithms of $2^{n} \leq \sum_{j = 0}^{L} C_{j}$ gives
\begin{equation*}
    n \leq \log_{2}\left(\sum_{j = 0}^{L} C_{j}\right) = L \log_{2}\varphi + O(1)
\end{equation*}
and hence
\begin{equation*}
    L \geq \frac{n}{\log_{2}\varphi} + O(1) \approx 1.440420n + O(1)
\end{equation*}
So any such code must incur an asymptotic expansion factor of at least $1.440420\ldots$.

% Note I.27
\noteblock{I.27}{Congruence languages.}

The key observation is that, when a binary digit $b \in \{0,1\}$ is appended to a word whose
current value is congruent to $r \pmod{7}$, the new value is congruent to
\begin{equation*}
    2r + b \pmod{7}.
\end{equation*}
So it is enough to remember the current remainder modulo $7$. This leads to the standard
remainder automaton for binary strings congruent to $2 \pmod{7}$:
\begin{itemize}
    \item \textbf{States}: A set $\mathcal{Q} = \{S_{0}, \dots, S_{6}\}$, where each state $S_{r}$ represents the current remainder $r$ modulo 7.
    \item \textbf{Initial State}: $S_{0}$ (representing the value 0).
    \item \textbf{Accepting State}: $\overline{Q} = \{S_{2}\}$.
    \item \textbf{Alphabet}: $\mathcal{A} = \{0, 1\}$.
\end{itemize}
Each transition follows the update rule $q_{r} \circ b = q_{(2r + b) \pmod{7}}$.
\begin{table}[H]
    \centering
    \caption{State Transition}
    \label{tab:flexible}
    \begin{tabularx}{\textwidth}{XXX}  % All center-aligned
        \toprule
        Current State (Remainder) & Input 0 ($2r \pmod{7}$) & Input 1 ($2r + 1 \pmod{7}$) \\
        \midrule
        $S_{0}$ & $S_{0}$ & $S_{1}$ \\
        $S_{1}$ & $S_{2}$ & $S_{3}$ \\
        $S_{2}$ & $S_{4}$ & $S_{5}$ \\
        $S_{3}$ & $S_{6}$ & $S_{0}$ \\
        $S_{4}$ & $S_{1}$ & $S_{2}$ \\
        $S_{5}$ & $S_{3}$ & $S_{4}$ \\
        $S_{6}$ & $S_{5}$ & $S_{6}$ \\
        \bottomrule
    \end{tabularx}
\end{table}
Any binary word is accepted exactly when the automaton finishes in state $S_{2}$, so the
language is A-regular.

Nothing here is special to modulus $7$ or to base $2$. For representations in base $b$, one
replaces the update rule by
\begin{equation*}
    r \mapsto br + d \pmod{m},
\end{equation*}
where $d$ is the next digit. Thus congruence conditions modulo $m$ are always recognized by
a finite automaton, and so are finite boolean combinations of such conditions.

% Note I.29
\noteblock{I.29}{The forward equations.}

Let $\mathcal{M}_{k}$ be the set of words that lead from the initial state $q_{0}$ to the state
$q_{k}$. A last-letter decomposition gives
\begin{equation*}
    \mathcal{M}_{k} \cong \llbracket k = 0 \rrbracket \cdot \{\epsilon\} +
        \sum_{j, \alpha: q_{j} \circ \alpha = q_{k}} (\mathcal{M}_{j} \times \{\alpha\})
\end{equation*}
where $\llbracket \cdot \rrbracket$ is the Iverson bracket. Indeed, a word reaching $q_{k}$ is either
the empty word when $k = 0$, or else some word reaching a predecessor state $q_{j}$,
followed by a final letter $\alpha$ that sends $q_{j}$ to $q_{k}$.

Translating directly to generating functions gives
\begin{equation*}
    M_{k}(z) = \delta_{k, 0} + z \sum_{j} M_{j}(z) T_{j, k}.
\end{equation*}
In vector form, if $\mathbf{M}(z)$ is the row vector $(M_{0}(z), \dots, M_{s}(z))$, then
\begin{equation*}
    \mathbf{M}(z) = \mathbf{e}_{0}^{T} + z\mathbf{M}(z)T,
\end{equation*}
where $\mathbf{e}_{0}^{T} = (1, 0, \dots, 0)$ is exactly the row vector denoted by $u$ in
Proposition~I.3. The forward equations count paths from the distinguished initial state
$q_{0}$ to each state $q_{k}$, so the length-$0$ contribution appears only at $k = 0$.
By contrast, the vector $v$ records which states are final, so it belongs to the backward
formulation. Once $\mathbf{M}(z)$ is known, the accepted language is recovered only at the end by
selecting the final states:
\begin{equation*}
    L(z) = \mathbf{M}(z)v.
\end{equation*}

% Note I.33.
\noteblock{I.33}{Waiting times in strings.}
Let $L \subseteq \mathrm{SEQ}\{a, b\}$ be a language, and let $\hat{\mathcal{L}}$ denote its
prefix language, namely the set of words in $L$ that have no strict prefix belonging to $L$.
For a random infinite binary string with $\mathbb{P}(a) = \mathbb{P}(b) = 1/2$, the quantity
$\hat{L}(1/2)$ is the probability that the string starts with a word of $\hat{\mathcal{L}}$,
while
\begin{equation*}
    \frac{1}{2} \hat{L}'(1/2)
\end{equation*}
gives the expected matched length contributed by the first word of the prefix language that
is encountered. In the regular case, this expectation is therefore rational.
This quantity should be read as a prefix-language statistic: it averages the length of the first
matched prefix, with value $0$ when no word of $\hat{\mathcal{L}}$ occurs at the beginning.

Let us test these principles on a deliberately tiny prefix language,
$\hat{\mathcal{L}} = \{aa, bbb\}$.
\begin{enumerate}
    \item Prefix property: $aa$ is not a prefix of $bbb$, and conversely, so the language is prefix-free.
    \item Generating function: the OGF is
        $\hat{L}(z) = z^{2} + z^{3}$.
    \item Probability: $\hat{L}(1/2)$ is the probability that a random infinite string starts with a word of $\hat{\mathcal{L}}$:
        $\hat{L}(1/2) = (1/2)^{2} + (1/2)^{3} = 1/4 + 1/8 = 3/8$.
    \item Expected matched length: $\frac{1}{2} \hat{L}'(1/2)$ gives the expected length contributed by the first matched word:
        $\hat{L}'(z) = 2z + 3z^{2} \implies \hat{L}'(1/2) = 1.75 \implies E = 0.875$.
    \item Unpacking the value:
        \begin{itemize}
            \item Word ``aa'': Probability $1/4$ of length 2 $\rightarrow$ Contribution: $2 \times 0.25 = 0.5$.
            \item Word ``bbb'': Probability $1/8$ of length 3 $\rightarrow$ Contribution: $3 \times 0.125 = 0.375$.
            \item No Match: Probability $5/8$ of length 0 $\rightarrow$ Contribution: $0 \times 0.625 = 0$.
        \end{itemize}
\end{enumerate}

% Note I.34.
\noteblock{I.34}{A probabilistic paradox on strings.}

Fix a binary pattern of length $k$. The generating function $T(z)$ for its first occurrence is:
\begin{equation*}
    T(z) = \frac{z^{k}}{z^{k} + (1 - mz)c(z)} = \frac{z^{k}}{z^{k} + (1 - 2z)c(z)}
\end{equation*}
The expected waiting time then follows immediately from the derivative at $z = 1/2$:
\begin{equation*}
    \left.z \frac{\dd}{\dd z} T(z)\right|_{z = 1/2} = 2^{k} c(1/2)
\end{equation*}
Unlike Note~I.33, this is not a prefix-language average with a possible value $0$; it is the
expected position of the first occurrence of a fixed factor inside the whole infinite sequence.
Thus the mean first-occurrence time is governed by the autocorrelation polynomial $c(z)$.
Patterns with stronger self-overlap have larger values of $c(1/2)$ and therefore tend to appear
later for the first time.

This already leads to a striking comparison. For $p = abb$, the only overlap is the trivial one,
so $c(z) = 1$ and the average first-occurrence time is
\begin{equation*}
    2^{3} c(1/2) = 8.
\end{equation*}
For $p' = aaa$, one has the proper self-overlaps of lengths $1$ and $2$, hence
\begin{equation*}
    c(z) = 1 + z + z^{2},
\end{equation*}
and therefore
\begin{equation*}
    2^{3} c(1/2) = 8\left(1 + \frac{1}{2} + \frac{1}{4}\right) = 14.
\end{equation*}
So the pattern $abb$ tends to occur sooner than $aaa$.

At first sight this seems paradoxical. All binary patterns of length $3$ have the same expected
number of occurrences in a long random word, yet their average first-occurrence times can be
quite different. The resolution is that $aaa$ overlaps with itself, so its occurrences tend to come
in clusters: once one copy appears, nearby copies become more likely. Those clusters are then
separated by longer barren stretches. By contrast, occurrences of $abb$ are more evenly spread.
There is therefore no contradiction between equal average counts and unequal average waiting
times.

\section{Tree structures}
% Note I.39.
\noteblock{I.39}{Unary–binary trees and Motzkin numbers.}

Unary--binary trees are rooted plane trees in which every node has either $0$, $1$, or $2$
children. Their symbolic specification is therefore
\begin{equation*}
    \mathcal{M} = \mathcal{Z} \times \mathrm{SEQ}_{\leq 2}(\mathcal{M}),
\end{equation*}
so the characteristic polynomial is
\begin{equation*}
    \phi(u) = 1 + u + u^{2}.
\end{equation*}
Consequently, the OGF satisfies
\begin{equation*}
    M(z) = z(1 + M(z) + M(z)^{2}),
\end{equation*}
that is,
\begin{equation*}
    zM(z)^{2} - (1 - z)M(z) + z = 0.
\end{equation*}
Solving the quadratic and choosing the branch with $M(0) = 0$ gives
\begin{equation*}
    M(z) = \frac{1 - z - \sqrt{1 - 2z - 3z^{2}}}{2z}.
\end{equation*}
Its expansion begins as
\begin{equation*}
    M(z) = z + z^{2} + 2z^{3} + 4z^{4} + 9z^{5} + 21z^{6} + 51z^{7} + \cdots.
\end{equation*}
The coefficients $M_{n} = [z^{n}]M(z)$ are the Motzkin numbers.

To extract them directly, we apply Lagrange inversion:
\begin{equation*}
    \begin{aligned}
        M_{n}
        &= \frac{1}{n} [u^{n - 1}] \phi(u)^{n} \\
        &= \frac{1}{n} [u^{n - 1}] (1 + u + u^{2})^{n} \\
        &= \frac{1}{n} [u^{n - 1}] \sum_{j} \binom{n}{j}(u + u^{2})^{j} \\
        &= \frac{1}{n} \sum_{j} \binom{n}{j} [u^{n - 1}] u^{j}(1 + u)^{j} \\
        &= \frac{1}{n} \sum_{j} \binom{n}{j} [u^{n - j - 1}] \sum_{l} \binom{j}{l} u^{l} \\
        &= \frac{1}{n} \sum_{j} \binom{n}{j} \binom{j}{n - j - 1} \\
        &= \frac{1}{n} \sum_{k} \binom{n}{n - k} \binom{n - k}{n - (n - k) - 1} \\
        &= \frac{1}{n} \sum_{k} \binom{n}{k} \binom{n - k}{k - 1}
    \end{aligned}
\end{equation*}
which is the explicit form quoted in the book.

% Note I.41.
\noteblock{I.41}{The Lagrangean form of Schröder’s GF.}

Only the last equality in the book needs a proof:
\begin{equation*}
    \frac{(-1)^{n - 1}}{n} \sum_{k}(-2)^{k} \binom{n}{k + 1} \binom{n + k - 1}{k} =
        \frac{1}{n} \sum_{k = 0}^{n - 2} \binom{2n - k - 2}{n - 1} \binom{n - 2}{k}
\end{equation*}
so it is enough to show
\begin{equation*}
    (-1)^{n - 1} \sum_{k = 0}^{n - 1}(-2)^{k} \binom{n}{k + 1} \binom{n + k - 1}{k} = \sum_{k = 0}^{n - 2} \binom{2n - k - 2}{n - 1} \binom{n - 2}{k}
\end{equation*}
Let the left-hand side be $L$ and the right-hand side be $R$.

Re-index the left sum by $j = n - 1 - k$. Using
$\binom{n}{k + 1} = \binom{n}{n - 1 - k}$ and
$\binom{n + k - 1}{k} = \binom{n + k - 1}{n - 1}$, we get
\begin{equation*}
    L = (-1)^{n - 1} \sum_{j = 0}^{n - 1}(-2)^{n - 1 - j} \binom{n}{j} \binom{2n - 2 - j}{n - 1} =
        2^{n - 1} \sum_{j = 0}^{n - 1} \left(-\frac{1}{2} \right)^{j} \binom{n}{j} \binom{2n - 2 - j}{n - 1}
\end{equation*}

Since $\binom{2n - 2 - j}{n - 1} = [z^{n - 1}] (1 + z)^{2n - 2 - j}$, coefficient extraction yields
\begin{equation*}
    \begin{aligned}
        L
        &= 2^{n - 1} \sum_{j = 0}^{n - 1} \left(-\frac{1}{2} \right)^{j} \binom{n}{j} [z^{n - 1}] (1 + z)^{2n - 2 - j} \\
        &= 2^{n - 1}[z^{n - 1}] \sum_{j = 0}^{n - 1} \left(-\frac{1}{2} \right)^{j} \binom{n}{j}(1 + z)^{2n - 2 - j} \\
        &= 2^{n - 1}[z^{n - 1}] (1 + z)^{2n - 2} \sum_{j = 0}^{n} \binom{n}{j} \left(-\frac{1}{2(1 + z)} \right)^{j} \\
        &= 2^{n - 1}[z^{n - 1}] (1 + z)^{2n - 2}\left(1 - \frac{1}{2(1 + z)} \right)^{n} \\
        &= 2^{n - 1}[z^{n - 1}] (1 + z)^{2n - 2} \frac{(1 + 2z)^{n}}{2^{n}(1 + z)^{n}} \\
        &= \frac{1}{2} [z^{n - 1}] (1 + z)^{n - 2}(1 + 2z)^{n}
    \end{aligned}
\end{equation*}
Here the upper limit has been extended from $n - 1$ to $n$; the added term has degree
$n - 2$ inside the coefficient extraction, so it contributes nothing to $[z^{n - 1}]$.
Expanding $(1 + 2z)^{n}$ now gives
\begin{equation*}
    \begin{aligned}
        L
        &= \frac{1}{2} \sum_{j = 0}^{n} \binom{n}{j} 2^{j} [z^{n - 1 - j}] (1 + z)^{n - 2} \\
        &= \frac{1}{2} \sum_{j = 0}^{n} \binom{n}{j} 2^{j} \binom{n - 2}{n - 1 - j}.
    \end{aligned}
\end{equation*}

Likewise, $\binom{2n - k - 2}{n - 1} = [z^{n - 1}] (1 + z)^{2n - k - 2}$, so
\begin{equation*}
    \begin{aligned}
        R
        &= \sum_{k = 0}^{n - 2} \binom{2n - k - 2}{n - 1} \binom{n - 2}{k} \\
        &= \sum_{k = 0}^{n - 2} [z^{n - 1}] (1 + z)^{2n - k - 2} \binom{n - 2}{k} \\
        &= [z^{n - 1}] \sum_{k = 0}^{n - 2}(1 + z)^{2n - k - 2} \binom{n - 2}{k} \\
        &= [z^{n - 1}] (1 + z)^{2n - 2} \sum_{k = 0}^{n - 2} \binom{n - 2}{k}(1 + z)^{-k} \\
        &= [z^{n - 1}] (1 + z)^{2n - 2}\left(1 + \frac{1}{1 + z} \right)^{n - 2} \\
        &= [z^{n - 1}] (1 + z)^{2n - 2} \frac{(2 + z)^{n - 2}}{(1 + z)^{n - 2}} \\
        &= [z^{n - 1}] (1 + z)^{n}(2 + z)^{n - 2}
    \end{aligned}
\end{equation*}
\begin{equation*}
    \begin{aligned}
        R
        &= \sum_{k = 0}^{n - 2} \binom{n - 2}{k} 2^{n - 2 - k} [z^{n - 1 - k}] (1 + z)^{n} \\
        &= \sum_{k = 0}^{n - 2} \binom{n - 2}{k} 2^{n - 2 - k} \binom{n}{k + 1}.
    \end{aligned}
\end{equation*}
Now set $j = n - 1 - k$. Then
\begin{equation*}
    R = \frac{1}{2} \sum_{j = 1}^{n - 1} \binom{n}{j} 2^{j} \binom{n - 2}{n - 1 - j}.
\end{equation*}
The missing endpoints $j = 0$ and $j = n$ contribute nothing, since
\[
    \binom{n - 2}{n - 1} = \binom{n - 2}{-1} = 0.
\]
Therefore
\begin{equation*}
    R = \frac{1}{2} \sum_{j = 0}^{n} \binom{n}{j} 2^{j} \binom{n - 2}{n - 1 - j} = L,
\end{equation*}
which is exactly the last equality claimed in the book.

% Note I.42.
\noteblock{I.42}{Faster determination of Schröder numbers.}
Start from the closed form
\begin{equation*}
    S(z) = \frac{1}{4}(1 + z - \sqrt{1 - 6z + z^{2}}).
\end{equation*}
Equivalently,
\begin{equation*}
    1 + z - 4S(z) = \sqrt{1 - 6z + z^{2}}.
\end{equation*}
Differentiating both sides gives
\begin{equation*}
    1 - 4S'(z) = \frac{z - 3}{\sqrt{1 - 6z + z^{2}}}.
\end{equation*}
Let $R(z) = 1 - 6z + z^{2}$. Multiplying the differentiated relation by $R(z)$ gives
\begin{equation*}
    R(z)(1 - 4S'(z)) = (z - 3)\sqrt{R(z)}.
\end{equation*}
Now use $\sqrt{R(z)} = 1 + z - 4S(z)$:
\begin{equation*}
    R(z)(1 - 4S'(z)) = (z - 3)(1 + z - 4S(z)).
\end{equation*}
Expanding gives
\begin{equation*}
    R(z) - 4R(z)S'(z) = (z - 3)(1 + z) - 4(z - 3)S(z).
\end{equation*}
Since $R(z) - (z - 3)(1 + z) = 4(1 - z)$, this rearranges to
\begin{equation*}
    R(z)S'(z) + (3 - z)S(z) + z - 1 = 0,
\end{equation*}
that is, the linear differential equation
\begin{equation*}
    (z^{2} - 6z + 1)S'(z) + (3 - z)S(z) + z - 1 = 0.
\end{equation*}

Write
\begin{equation*}
    S(z) = \sum_{n \geq 1} S_{n} z^{n},
    \qquad
    S'(z) = \sum_{n \geq 0} (n + 1)S_{n + 1} z^{n}.
\end{equation*}
Substituting into the differential equation and extracting the coefficient of $z^{n}$ for
$n \geq 2$ gives
\begin{equation*}
    (n - 1)S_{n - 1} - 6nS_{n} + (n + 1)S_{n + 1} + 3S_{n} - S_{n - 1} = 0,
\end{equation*}
that is,
\begin{equation*}
    (n - 2)S_{n - 1} - 3(2n - 1)S_{n} + (n + 1)S_{n + 1} = 0.
\end{equation*}
Replacing $n$ by $n + 1$ yields the recurrence in the book:
\begin{equation*}
    (n + 2)S_{n + 2} - 3(2n + 1)S_{n + 1} + (n - 1)S_{n} = 0,
    \qquad n \geq 1.
\end{equation*}

The underlying mechanism is more general. Whenever an algebraic generating function is given
in the form
\begin{equation*}
    p(z)F(z) + q(z) = \sqrt[k]{r(z)},
\end{equation*}
one can differentiate, eliminate the radical, and often obtain a linear differential equation
for $F(z)$. Coefficient extraction then turns that differential equation into a recurrence for the
coefficients. In the present case, this general holonomic principle leads to the three-term
recurrence above.

% Note I.43.
\noteblock{I.43}{Fast determination of the Cayley–Pólya numbers.}
For rooted non-plane trees, Cayley's functional equation is
\begin{equation*}
    H(z) = z \exp\left(H(z) + \frac{1}{2} H(z^{2}) + \cdots\right)
\end{equation*}
or, equivalently,
\begin{equation*}
    H(z) = z \exp\left(\sum_{k \geq 1} \frac{1}{k}H(z^{k})\right).
\end{equation*}
Since no closed form is available, the useful point is that logarithmic differentiation still
gives an efficient recurrence. Taking the logarithm gives
\begin{equation*}
    \log H(z) = \log z + \sum_{k \geq 1} \frac{1}{k}H(z^{k}),
\end{equation*}
and differentiating yields
\begin{equation*}
    \frac{H'(z)}{H(z)}
        = \frac{1}{z} + \sum_{k \geq 1} z^{k - 1}H'(z^{k}).
\end{equation*}
Multiplying by $zH(z)$ clears the denominator:
\begin{equation*}
    zH'(z) = H(z)\left(1 + \sum_{k \geq 1} z^{k}H'(z^{k})\right).
\end{equation*}

Now write $H(z) = \sum_{n \geq 1}H_{n}z^{n}$. The inner sum has the divisor form
\begin{equation*}
    \sum_{k \geq 1} z^{k}H'(z^{k})
        = \sum_{k \geq 1}\sum_{m \geq 1} mH_{m}z^{km}
        = \sum_{\ell \geq 1}\left(\sum_{d|\ell} dH_{d}\right)z^{\ell}.
\end{equation*}
Thus, if
\begin{equation*}
    a_{\ell} = \sum_{d|\ell} dH_{d},
\end{equation*}
then coefficient extraction in
\[
    zH'(z) = H(z)\left(1 + \sum_{\ell \geq 1} a_{\ell}z^{\ell}\right)
\]
gives, for $n \geq 2$,
\begin{equation*}
    (n - 1)H_{n} = \sum_{l = 1}^{n - 1} H_{n - l}\left(\sum_{d|l} d H_{d}\right)
\end{equation*}
or, equivalently,
\begin{equation*}
    H_{n}
    = \frac{1}{n - 1} \sum_{l = 1}^{n - 1} H_{n - l}\left(\sum_{d|l} d H_{d}\right)
    = \frac{1}{n - 1} \sum_{l = 1}^{n - 1} a_{l}H_{n - l}.
\end{equation*}
This determines $H_{n}$ from earlier values only. Computing the divisor sums $a_l$ as the
values are needed gives a polynomial-time algorithm for the Cayley--Polya numbers.

To see the recurrence at work, start from the base case $H_{1} = 1$, corresponding to the
one-node tree.
\begin{itemize}
    \item To find $H_{2}$, we set $n = 2$. The summation runs for $l = 1$. The only divisor of $1$ is $1$.
        $a_{1} = \sum_{d|1} d H_{d} = 1 \cdot H_{1} = 1 \implies
        H_{2} = \frac{1}{2 - 1}(H_{2 - 1} \cdot a_{1}) = 1$.
    \item For $H_{3}$, the summation runs for $l = 1$ and $l = 2$. The divisors of $2$ are $1$ and $2$.
        $a_{2} = \sum_{d|2} d H_{d} = 1 \cdot H_{1} + 2 \cdot H_{2} = 3 \implies
        H_{3} = \frac{1}{3 - 1}(H_{3 - 1} \cdot a_{1} + H_{3 - 2} \cdot a_{2}) = 2$.
    \item For $H_{4}$, the summation runs for $l = 1, 2, 3$. The divisors of $3$ are $1$ and $3$.
        $a_{3} = \sum_{d|3} d H_{d} = 1 \cdot H_{1} + 3 \cdot H_{3} = 7 \implies
        H_{4} = \frac{1}{4 - 1}(H_{4 - 1} \cdot a_{1} + H_{4 - 2} \cdot a_{2} + H_{4 - 3} \cdot a_{3}) = 4$.
\end{itemize}

% Note I.50.
\noteblock{I.50}{Excursions, bridges, and meanders.}

We use the step set $\{-1, +1\}$. A meander is a walk whose partial sums are always
non-negative, a bridge is a walk whose final sum is $0$, and an excursion is both a meander
and a bridge. Thus excursions are exactly Dyck paths.

A meander ending at altitude $k$ can be read as $k$ up-steps that create new floors, with Dyck
excursions $\mathcal{D}$ inserted before, between, and after these up-steps:
\begin{equation*}
    \mathcal{M}
        = \mathcal{D}
            + (\mathcal{D} \times \mathcal{Z} \times \mathcal{D})
            + (\mathcal{D} \times \mathcal{Z} \times \mathcal{D} \times \mathcal{Z} \times \mathcal{D})
            + \cdots
        = \mathcal{D} \times \sum_{k = 0}^{\infty}(\mathcal{Z} \times \mathcal{D})^{k}
\end{equation*}
Therefore
\begin{equation*}
    M(z) = \frac{D(z)}{1 - zD(z)}.
\end{equation*}

A bridge is a sequence of primitive excursions. Each primitive excursion leaves 0, avoids returning in the middle, and comes back only at the end.
\begin{itemize}
    \item \textbf{Going Above:} Starts with +1, ends with -1, with a Dyck path in between: $\mathcal{Z} \times \mathcal{D} \times \mathcal{Z}$
    \item \textbf{Going Down:} Starts with -1, ends with 1, with a mirrored Dyck path in between: $\mathcal{Z} \times \mathcal{D} \times \mathcal{Z}$
\end{itemize}
\begin{equation*}
    \mathcal{B}
        = \mathrm{S}\textsc{eq}(\mathcal{Z} \times \mathcal{D} \times \mathcal{Z} + \mathcal{Z} \times \mathcal{D} \times \mathcal{Z})
        = \mathrm{S}\textsc{eq}(2 \times \mathcal{Z}^{2} \times \mathcal{D})
\end{equation*}
Thus
\begin{equation*}
    B(z) = \frac{1}{1 - 2z^{2}D(z)}.
\end{equation*}

The closed forms of these coefficients are classical. For meanders, let $N(n, k)$ be the
number of unrestricted paths of length $n$ ending at altitude $k$.
The reflection principle does the main work: paths that ever hit $-1$ and end at $k \geq 0$ are
in bijection with unrestricted paths ending at $-k - 2$.
So the number of meanders ending at $k$ is $N(n, k) - N(n, k + 2)$, and summing over $k \geq 0$ gives the telescoping identity
\begin{equation*}
    M_{n} = \sum_{k \geq 0} [N(n, k) - N(n, k + 2)]
\end{equation*}
\begin{itemize}
    \item $n$ is even ($n = 2m$)
        $M_{n} = N(n, 0) = N(2m, 0) = \binom{2m}{\frac{2m + 0}{2}} = \binom{2m}{m} = \binom{n}{\lfloor n/2 \rfloor}$
    \item $n$ is odd ($n = 2m + 1$)
        $M_{n} = N(n, 1) = N(2m + 1, 1) = \binom{2m + 1}{m + 1}
            = \binom{2m + 1}{m} = \binom{n}{\lfloor n/2 \rfloor}$
\end{itemize}
Hence
\begin{equation*}
    M_{n} = \binom{n}{\lfloor n/2 \rfloor}.
\end{equation*}
For bridges, the final altitude is $0$, so only even lengths contribute, and an unrestricted
choice of which $n$ of the $2n$ steps are up-steps gives
\begin{equation*}
    B_{2n} = \binom{2n}{n}.
\end{equation*}

% Example I.17.
\exampleblock{I.17}{The complexity of boolean functions.}

Let $\Omega$ be the set of all boolean functions in $m$ variables. Then
\begin{equation*}
    \lVert \Omega \rVert = 2^{2^{m}}.
\end{equation*}
As descriptions, use binary expression trees whose internal nodes are labelled by
$\vee$ or $\wedge$, and whose external nodes are labelled by one of the $2m$ literals
$x_{1}, \neg x_{1}, \dots, x_{m}, \neg x_{m}$. The size of a description is the number
of internal nodes, namely the number of logical connectives.

If a description has size $n$, its underlying full binary tree has $n$ internal nodes and
$n + 1$ external nodes. There are
\begin{equation*}
    \frac{1}{n + 1}\binom{2n}{n}
\end{equation*}
such tree shapes. The internal nodes have $2^{n}$ possible connective labels, and the leaves
have $(2m)^{n + 1}$ possible literal labels. Hence
\begin{equation*}
    D_{n} = \frac{1}{n + 1}\binom{2n}{n}\,2^{n}(2m)^{n + 1}.
\end{equation*}
Using the Catalan estimate
\[
    \frac{1}{n + 1}\binom{2n}{n} = O(4^{n}n^{-3/2}),
\]
we get, for fixed $m$,
\begin{equation*}
    D_{n} = O(16^{n}m^{n}n^{-3/2}).
\end{equation*}
Equivalently, keeping the dependence on $m$ visible,
\begin{equation*}
    D_{n} = O(m(16m)^{n}n^{-3/2}).
\end{equation*}

The counting argument is simple. If
\begin{equation*}
    \sum_{j = 0}^{\nu} D_{j} < \lVert \Omega \rVert,
\end{equation*}
then descriptions of size at most $\nu$ cannot represent all boolean functions. If the left-hand
side is $o(\lVert \Omega \rVert)$, then the fraction of functions of complexity at most $\nu$
tends to $0$.

Here is the elementary estimate behind the partial sum. If $c > 1$, $\alpha > 0$, and
$a_{j} = c^{j}j^{-\alpha}$, then
\begin{equation*}
    \sum_{j = 1}^{\nu} a_{j} = O(c^{\nu}\nu^{-\alpha}).
\end{equation*}
Indeed, split the sum at $\lfloor \nu/2 \rfloor$. For the early part, $j^{-\alpha} \leq 1$ and
the geometric sum gives
\begin{equation*}
    \sum_{j \leq \nu/2} c^{j}j^{-\alpha}
        \leq \sum_{j \leq \nu/2} c^{j}
        = O(c^{\nu/2}).
\end{equation*}
This is negligible on the target scale, since
\begin{equation*}
    \frac{c^{\nu/2}}{c^{\nu}\nu^{-\alpha}}
        = \frac{\nu^{\alpha}}{c^{\nu/2}} \longrightarrow 0.
\end{equation*}
For the tail, $j \geq \nu/2$ implies $j^{-\alpha} \leq 2^{\alpha}\nu^{-\alpha}$, whence
\begin{equation*}
    \sum_{\nu/2 < j \leq \nu} c^{j}j^{-\alpha}
        \leq 2^{\alpha}\nu^{-\alpha}\sum_{\nu/2 < j \leq \nu} c^{j}
        = O(c^{\nu}\nu^{-\alpha}).
\end{equation*}
Combining both parts proves the estimate.

Apply this with $c = 16m$ and $\alpha = 3/2$. With the explicit factor of $m$, there are
constants $C$ and $j_{0}$ such that, for $j \geq j_{0}$,
\[
    D_{j} \leq Cm(16m)^{j}j^{-3/2}.
\]
The finitely many terms with $j < j_{0}$ only contribute a constant, and the estimate above
controls the remaining tail. Therefore
\begin{equation*}
    \sum_{j = 0}^{\nu}D_{j} = O(m16^{\nu}m^{\nu}\nu^{-3/2})
\end{equation*}
When $m$ is fixed, the leading factor can of course be absorbed into the constant. Now choose
\begin{equation*}
    \nu(m) = \frac{2^{m}}{4 + \log_{2}m}.
\end{equation*}
Then
\begin{equation*}
    \nu(m)\log_{2}(16m)
        = \frac{2^{m}}{4 + \log_{2}m}(4 + \log_{2}m)
        = 2^{m},
\end{equation*}
so
\begin{equation*}
    16^{\nu(m)}m^{\nu(m)} = 2^{2^{m}} = \lVert \Omega \rVert.
\end{equation*}
Consequently,
\begin{equation*}
    \sum_{j = 0}^{\nu(m)}D_{j}
        = O(m\lVert \Omega \rVert\,\nu(m)^{-3/2})
        = o(\lVert \Omega \rVert).
\end{equation*}
Indeed, the ratio is $O(m\nu(m)^{-3/2})$, which still tends to $0$.
Thus a fraction tending to $1$ of boolean functions in $m$ variables have tree complexity at
least
\begin{equation*}
    \frac{2^{m}}{4 + \log_{2}m}.
\end{equation*}

This lower bound is close to the elementary upper bound. For $m = 1$, the four boolean
functions are represented by
\begin{equation*}
    0 \equiv (x_{1}\wedge \neg x_{1}), \qquad
    1 \equiv (x_{1}\vee \neg x_{1}), \qquad
    x_{1}, \qquad \neg x_{1}.
\end{equation*}
For $m > 1$, use the binary decision tree decomposition
\begin{equation*}
    f(x_{1}, \dots, x_{m})
        =
        (\neg x_{m}\wedge f(x_{1}, \dots, x_{m - 1}, 0))
        \vee
        (x_{m}\wedge f(x_{1}, \dots, x_{m - 1}, 1)).
\end{equation*}
This reduces an $m$-variable function to two $(m - 1)$-variable functions and uses three new
connectives, so induction gives an upper bound of $2^{m + 1} - 3$ connectives. Hence most
boolean functions have tree complexity of order $2^{m}/\log m$, while every one has
complexity $O(2^{m})$.

\section{Additional constructions}
% Note I.62.
\noteblock{I.62}{Sequences without repeated components.}

A sequence without repeated components is an ordered arrangement of distinct objects drawn
from a base class $\mathcal{B}$. The natural route is to build an unordered set of distinct
components first, and then order it afterward.

Introduce a marker $u$ for the number of selected components. The bivariate generating
function for the power-set construction is
\begin{equation*}
    A(z, u)
        = \prod_{\beta \in \mathcal{B}}\left(1 + uz^{|\beta|}\right)
        = \exp\left(\sum_{j \geq 1}(-1)^{j - 1} \frac{u^{j}}{j}B(z^{j})\right)
\end{equation*}
where $B(z)$ is the OGF of $\mathcal{B}$. Write
\begin{equation*}
    A(z, u) = \sum_{k \geq 0} f^{\langle k \rangle}(z)u^{k}.
\end{equation*}
Then $f^{\langle k \rangle}(z)$ counts sets with exactly $k$ distinct components. To turn such
sets into sequences, multiply by $k!$, since a set of $k$ distinct objects has $k!$ possible
orderings.

Using the Eulerian integral
\[
    k! = \int_{0}^{\infty} e^{-t}t^{k}\,\dd t,
\]
with $t$ as the integration variable, the OGF of sequences without repeated components is
\begin{equation*}
    \begin{aligned}
        \mathcal{S}_{\text{distinct}}(z)
            &= \sum_{k \geq 0} k! f^{\langle k \rangle}(z) \\
            &= \int_{0}^{\infty} e^{-t}\sum_{k \geq 0} f^{\langle k \rangle}(z)t^{k}\,\dd t \\
            &= \int_{0}^{\infty} A(z, t)e^{-t}\,\dd t \\
            &= \int_{0}^{\infty}
                \exp\left(\sum_{j \geq 1}(-1)^{j - 1}\frac{t^{j}}{j}B(z^{j})\right)e^{-t}\,\dd t.
    \end{aligned}
\end{equation*}
This is the formula displayed in the book; the integral is formal, serving to implement the
factor $k!$ that orders a $k$-element set.

\section{Perspective}
